---
title: "Census Data"
author: " "
date: '2023-09-08'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load packages and functions

```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac")
```

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("67608da2d7728d1edc9cb636b83bb764101fe726", overwrite = TRUE)
```

```{r setup_packages, warning = FALSE, message = FALSE}
# Create a vector of years
years <- c(2017, 2018, 2019, 2020, 2021, 2022)

# Loop through the years to load variables
variables_list <- lapply(years, function(year) {
  load_variables(year, "acs5", cache = TRUE)
})

# Combine the variables from different years into a single data frame
all_variables <- do.call(rbind, variables_list)
```


### 2017 Census Data

```{r results='hide'}
tracts17 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2017, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2018 Census Data

```{r results='hide'}
tracts18 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2018, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2018") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2019 Census Data

```{r results='hide'}
tracts19 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2019, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2019") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2020 Census Data

```{r results='hide'}
tracts20 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2020, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2021 Census Data

```{r results='hide'}
tracts21 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2021, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2021") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2022 Census Data

```{r results='hide'}
tracts22 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2022, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2022") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### Combining Census Data from 2017 - 2022

```{r}

allTracts <- rbind(tracts17,tracts18,tracts19,tracts20,tracts21,tracts22)
```


### Relating CATS Stops and Tracts
#### Wrangling Charlotte Transit Open Data

```{r}

routes <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/Bus_Routes.geojson")

hlt <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/Bus_Stops_With_Frequency_HLT.geojson")

CATSstop <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/CATS_BusStops.shp")
```


```{r}

stopBuffer <- st_buffer(CATSstop, 2640)

stopUnion <- st_union(st_buffer(CATSstop, 2640))

CATSBuffer <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```


```{r}
ggplot() +
  geom_sf(data=CATSBuffer) +
  geom_sf(data=CATSstop, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 1") +
  mapTheme()
```


#### Using the `sf` Package for Spatial operations

```{r}
buffer <- filter(CATSBuffers, Legend=="Unioned Buffer")

buffer <- st_transform(buffer, st_crs(tracts17))
```


#### Spatial Intersection of tracts and buffers

##### Spatial Clipping with `st_intersection()` on polygons

```{r}
clip <- 
  st_intersection(buffer, tracts17) %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Clip")
```


##### Spatial selection

```{r spatialSelection}

# Do a spatial selection to see which tracts touch the buffer

# approach #1: sub-setting a spatial object with a spatial object using the '[' brackets.
selection1 <- 
  tracts17[buffer,] %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Spatial Selection")

```


##### Spatial intersection with with `st_centroid()` on polygon centroids

```{r}
selectCentroids <-
  st_centroid(tracts17)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts17, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Select by Centroids")
```

##### Plot

```{r}
intersections <- rbind(clip, selection1, selectCentroids)

ggplot() +
  geom_sf(data=intersections, aes(fill = Workwithbus)) +
  geom_sf(data=CATSstop, show.legend = "point", size = 0.7) +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

