---
title: "Combined Markdown"
author: "Teresa Chang, Roshini Ganesh, Jingyi Li, Jonathan Manurung"
date: "2024-02-27"
output: html_document
---

```{r setup, include=FALSE}
# You can set some global options for knitting chunks

knitr::opts_chunk$set(echo = TRUE)

# Load some libraries

library(tidyverse)
library(sf)
library(spdep)
library(caret)
library(ckanr)
library(FNN)
library(grid)
library(gridExtra)
library(ggcorrplot) # plot correlation plot
library(corrr)      # another way to plot correlation plot
library(kableExtra)
library(jtools)     # for regression model plots
library(ggstance) # to support jtools plots
library(ggpubr)    # plotting R^2 value on ggplot point scatter
library(broom.mixed) # needed for effects plots
library(tidycensus)
library(kableExtra)
library(stargazer)
library(readxl)
library(zoo)
library(cowplot)
library(patchwork)

# functions and data directory

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c")
palette7 <- c("white","#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c", "black")
red_palette <- c("#FFD3D3", "#FFA8A8", "#FF7D7D", "#FF5252", "#FF0000")
positive_palette <- c("darkgreen", "green", "yellow", "#FFAB00", "red")
negative_palette <- c("blue", "lightblue", "grey80", "grey60", "grey40")

positive_paletteaccent <- c("#006d2c", "#2ca25f", "#FFCB62", "#FFAB00", "#C60500")
negative_paletteaccent <- c("#0077B6", "#7ACFD3", "grey80", "grey60", "grey40")

palettebase5 <- c("#edf8fb","#7ACFD3","#3AA8AE","#2D8388","#005357")
palettebase7 <- c("#edf8fb","#7ACFD3","#54C1C6","#3AA8AE","#2D8388","#016F74","#005357")
palettebase <- c("#7ACFD3","#67C8C8","#54C1C6","#41BAC0","#3AA8AE","#33969B","#2D8388","#016F74","#005357")
paletteaccent <- c("#46AFB4","#249DA2","#01939A","#017278","#005A5E","#FFCB62","#FFBD37","#FFAB00","#C68500","#9C6900","#FF6662","#FF3C37","#FF0700","#C60500","#9C0400")

red_paletteaccent <- c("#FFD3D3","#FFA8A8","#FF0700","#C60500","#9C0400")
yellow_paletteaccent <- c("#FFE6A7","#FFBD37","#FFAB00","#C68500","#9C6900")

census_api_key("16325a5baedf1c986a182c3321f90022be22845a", overwrite = TRUE , install = TRUE)
```

## Read Data

```{r cars}
stops<-read_csv("AllDat.csv")
bus_routes <- st_read("Data/Bus_Routes.geojson")%>%
  st_transform('EPSG:3358')
school <- st_read("Data/Schools.geojson")%>%
  st_transform('EPSG:3358')

parks <- st_read("Data/Parks.geojson")%>%
  st_transform('EPSG:3358')

groceries <-
  st_read("Data/Grocery_Stores.geojson")%>%
  st_transform('EPSG:3358')

shoppingcen <-
  st_read("Data/Existing_Shopping_Centers.geojson")%>%
  st_transform('EPSG:3358')

policeoffice <-
  st_read("Data/CMPD_Police_Division_Offices.geojson") %>%
  st_transform('EPSG:3358')

# Create a vector of years
years <- c(2017, 2018, 2019, 2020, 2021, 2022)

# Loop through the years to load variables
variables_list <- lapply(years, function(year) {
  load_variables(year, "acs5", cache = TRUE)
})

# Combine the variables from different years into a single data frame
all_variables <- do.call(rbind, variables_list)

hlt <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.geojson")

# 2017 census
dat2017 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2017, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2018
dat2018 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2018, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2018") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

#census 2019
dat2019 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2019, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2019") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2020
dat2020 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2020, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale)

# census 2021
dat2021 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2021, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2021") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2022
dat2022 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2022, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2022") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```

## Data Wrangling

```{r pressure, echo=FALSE}
#feature engineering
school.sf <-
  school %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

parks.sf <-
  parks %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

groceries.sf <-
  groceries %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

shoppingcen.sf <-
  shoppingcen %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

policeoffice.sf <-
  policeoffice %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

silver_stops <- st_read("Data/LYNX_Silver_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

silver_routes <- st_read("Data/LYNX_Silver_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')

blue_stops <-
  st_read("Data/LYNX_Blue_Line_Stations.geojson")%>%
  st_transform('EPSG:3358')

blue_routes <-
  st_read("Data/LYNX_Blue_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

gold_stops <-
  st_read("Data/LYNX_Gold_Line_Stops.geojson")%>%
  st_transform('EPSG:3358')

gold_routes <-
  st_read("Data/LYNX_Gold_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

red_stops <-
  st_read("Data/LYNX_Red_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

red_routes <-
  st_read("Data/LYNX_Red_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')


allTracts <- rbind(
  dat2017,
  dat2018,
  dat2019,
  dat2020,
  dat2021,
  dat2022)
sf_allTracts <- st_as_sf(allTracts, wkt = "geometry")

bus_data <- stops %>% mutate(`Stop ID` = sprintf("%05s", `Stop ID`))
busstop_sf <- st_as_sf(bus_data, coords = c("Longitude", "Latitude"), crs = 4326)
catsstop_sf <- st_as_sf(stops, coords = c("Longitude", "Latitude"), crs = 2264)  # Assuming NAD 83/North Carolina (ftUS)
#Reproject catsstop_sf to WGS 84 (EPSG:4326)
catsstop_sf <- st_transform(catsstop_sf, 4326)
busstop_sf <- st_transform(busstop_sf, 3358)
catsstop_sf <- st_transform(catsstop_sf, 3358)
routes_sf <- st_transform(bus_routes, 3358)
allTracts_sf <- st_transform(allTracts, 3358)

```


```{r Board, warning = FALSE, eval = FALSE, results='hide'}
# Create an empty list to store subset dataframes
subset_list <- list()

# Loop through unique route numbers
for (route_num in unique(routes_sf$ROUTE_NUM)) {
  # Subset the routes for the current route
  subset_routes <- routes_sf[routes_sf$ROUTE_NUM == route_num, ]
  
  # Subset the stops for the current route
  subset_stops <- catsstop_sf[grepl(paste("\\b", route_num, "\\b", sep = ""), catsstop_sf$routes), ]
  
  # Combine routes and stops for the current route
  combined_subset <- bind_rows(subset_routes, subset_stops)
  
  # Add the combined subset to the list
  subset_list[[as.character(route_num)]] <- combined_subset
}
```

## APC data

```{r, warning = FALSE, eval = FALSE, results='hide'}
ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Board), size = .5) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Bus Stops", color = "Board Value")

ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Alight), size = .5) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Bus Stops", color = "Alight Value")
```

## Pre Covid vs. Post Covid

```{r}
# threshold: date of covid bgan
threshold_date <- as.Date("2020-03-01") 

stops_data <- stops %>%
  mutate(Numeric_Month = match(Month, month.name)) %>%
  mutate(Readable_Date = as.yearmon(paste(Year, sprintf("%02d", Numeric_Month)), "%Y %m")) %>%
  mutate(Covid_Status = ifelse(as.Date(Readable_Date) >= threshold_date, "Post", "Pre"))%>%
  rename(Stop_Name = `Stop Name`)%>%
  rename(Stop_ID = `Stop ID`)

# Calculate the increase/decrease
average_board <- stops_data %>%
  group_by(Stop_Name, Numeric_Month, Covid_Status) %>%
  summarize(Avg_Board = mean(Avg_Board, na.rm = TRUE))

average_board_pivoted <- pivot_wider(average_board, names_from = Covid_Status, values_from = Avg_Board)

average_board_pivoted <- average_board_pivoted %>%
  mutate(Difference = (Post - Pre)/Pre *100 ) %>%
  select(Stop_Name, Numeric_Month, Pre, Post, Difference)

average_board_pivoted <- na.omit(average_board_pivoted)
# join the longitude and latitude data back and select the first one
average_board_pivoted <- average_board_pivoted %>%
  left_join(stops_data %>% select(Stop_Name, Longitude, Latitude) %>% distinct(Stop_Name, .keep_all = TRUE), by = "Stop_Name")

# Adjust the month as needed
selected_month <- 1  
selected_data <- average_board_pivoted %>%
  filter(Numeric_Month == selected_month)

stops_sf <- st_as_sf(selected_data, coords = c("Longitude", "Latitude"), crs = 4326)
stops_sf <- st_transform(stops_sf, crs = 3358)  # NAD83 North Carolina

```

### All stops from 2017\~2023

```{r, fig.height=8, fig.width=16}
palettebase5_rev <- rev(palettebase5)

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = stops_sf , aes(colour = q5(Difference)), 
          show.legend = "point", size = 1, alpha= 0.5) +
  scale_colour_manual(values = palettebase5_rev,
                   labels=qBr(stops_sf ,"Difference"),
                   name="Quintile\nBreaks (%)") +
  labs(title = paste("Difference (%) - Month:", selected_month),
       fill = "Difference (Post - Pre)") +
  mapTheme()

filtered_stops_sf <- stops_sf[stops_sf$Difference != -100, ]

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = filtered_stops_sf , aes(colour = q5(Difference)), 
          show.legend = "point", size = 1, alpha= 0.5) +
  scale_colour_manual(values = palettebase5_rev,
                   labels=qBr(filtered_stops_sf ,"Difference"),
                   name="Quintile\nBreaks (%)") +
  labs(title = paste("Difference (%) (-100% excluded)   - Month:", selected_month),
       fill = "Difference (Post - Pre)") +
  mapTheme()
```

### Increase vs Decrease

```{r, fig.height=8, fig.width=16}
stops_sf$PositiveDifference <- stops_sf$Difference > 0
positive_stops_sf <- stops_sf[stops_sf$PositiveDifference,]
negative_stops_sf <- stops_sf[!stops_sf$PositiveDifference,]

yellow_paletteaccent_rev <- rev(yellow_paletteaccent)

# positive and negative differences
positive_plot <- ggplot(positive_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = red_paletteaccent,
                     labels = qBr(positive_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Positive Increase (%) - Month:", selected_month),
       fill = "Positive Difference (Post - Pre)") +
  mapTheme()

negative_plot <- ggplot(negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()

combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)
print(combined_plot)

# Exclude -100%
filtered_negative_stops_sf <- negative_stops_sf[negative_stops_sf$Difference != -100, ]

negative_plot <- ggplot(filtered_negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(filtered_negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) (-100% excluded) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()
combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)

print(combined_plot)
```

### Top 50 underperforming stops from 2017 to 2023

```{r, fig.height=8, fig.width=16}
# filter underperforming stops
underperforming_stops <- stops_data[(stops_data$Board + stops_data$Alight) <= 10, ]
underperformance_counts <- table(underperforming_stops$Stop_ID)


top_50_stops <- names(sort(underperformance_counts, decreasing = TRUE)[1:50])
underperforming_top_50 <- underperforming_stops[underperforming_stops$Stop_ID %in% top_50_stops, ]
underperforming_top_50 <- underperforming_top_50 %>%
  mutate(Total_Board_Alight = Board + Alight)

# Order
underperforming_top_50$Stop_ID <- factor(underperforming_top_50$Stop_ID, levels = rev(top_50_stops))

stacked_bar_plot <- ggplot(underperforming_top_50, aes(x = Stop_ID, fill = factor(Year))) +
  geom_bar(stat = "count", position = "stack", color = "darkgrey" ) +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = palettebase7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels
print(stacked_bar_plot) 

underperforming_top_50_all_sf <- st_as_sf(underperforming_top_50, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_all_sf <- st_transform(underperforming_top_50_all_sf, crs = 3358)  # NAD83 North Carolina


map_plot_top_50 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_all_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_all_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts (2017 - 2023)",
       fill = "Avg_Board + Avg_Alight") +
  mapTheme()
print(map_plot_top_50) 

combined_plot <- stacked_bar_plot + map_plot_top_50
print(combined_plot)
```

### Top 50 underperforming stops in 2022

```{r, fig.height=8, fig.width=16}
# Underperforming Stops in 2022
underperforming_stops_2022 <- stops_data[(stops_data$Board + stops_data$Alight) <= 10 & stops_data$Year == 2022, ]
underperformance_counts_2022 <- table(underperforming_stops_2022$Stop_ID)

# Top 30 Stops for 2023
top_50_stops_2022 <- names(sort(underperformance_counts_2022, decreasing = TRUE)[1:50])
underperforming_top_50_2022 <- underperforming_stops_2022[underperforming_stops_2022$Stop_ID %in% top_50_stops_2022, ]


# Order
underperforming_top_50_2022$Stop_ID <- factor(underperforming_top_50_2022$Stop_ID, levels = rev(top_50_stops_2022))

bar_plot_top_50_2022 <- ggplot(underperforming_top_50_2022, aes(x = Stop_ID, fill = "2022")) +
  geom_bar(stat = "count", color = "black") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = c("2022" = "#7ACFD3"), name = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
print(bar_plot_top_50_2022)

underperforming_top_50_2022 <- underperforming_top_50_2022 %>%
  mutate(Total_Board_Alight = Board + Alight)


underperforming_top_50_sf <- st_as_sf(underperforming_top_50_2022, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_sf <- st_transform(underperforming_top_50_sf, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_2022 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts in 2022",
       fill = "Total_Board_Alight") +
  mapTheme()
print(map_plot_top_50_2022)

combined_plot <- bar_plot_top_50_2022 + map_plot_top_50_2022
print(combined_plot)
```

### 2017\~2023 vs 2023

```{r, fig.height=10, fig.width=8}
print(map_plot_top_50) 
print(map_plot_top_50_2022)

print(stacked_bar_plot) 
print(bar_plot_top_50_2022)
```

### Existing and proposed transit lines and stations

```{r}
transit <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, aes(color = "Bus Route"), size = 5, alpha = 0.5) +
  geom_sf(data = blue_routes, aes(color = "Blue Line"), size = 8) +
  geom_sf(data = blue_stops, aes(color = "Blue Line"), size = 2) +
  geom_sf(data = gold_routes, aes(color = "Gold Line"), size = 8) +
  geom_sf(data = gold_stops, aes(color = "Gold Line"), size = 2) +
  geom_sf(data = silver_routes, aes(color = "Silver Line"), size = 8, linetype = "dashed") +
  geom_sf(data = silver_stops, aes(color = "Silver Line"), size = 2, shape = 21) +
  geom_sf(data = red_routes, aes(color = "Red Line"), size = 8, linetype = "dashed") +
  geom_sf(data = red_stops, aes(color = "Red Line"), size = 2, shape = 21) +
  
  labs(title = "Existing and Proposed Transit Lines and Stops") +
    scale_color_manual(name = "Lines",
                     values = c("Bus Route" = "darkgrey",
                                "Blue Line" = "#4682B4", 
                                "Gold Line" = "#FFD700", 
                                "Silver Line" = "purple", 
                                "Red Line" = "#FF6347"),
                     labels = c("Blue Line","Bus Routes", "Gold Line", "Proposed Red Line", "Proposed Silver Line")) +

  mapTheme()

print(transit)

```

### Density of Amenities

```{r, fig.height=8, fig.width=16}
A <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = school, color = "darkgrey", size = .5, alpha = 0.3) +
  stat_density2d(data = data.frame(st_coordinates(school.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFE6A7", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Schools") +
  mapTheme()

B <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = parks, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(parks.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#7ACFD3", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Parks") +
  mapTheme()

C <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = groceries, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(groceries.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFAB00", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Grocery Stores") +
  mapTheme()

D <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = policeoffice, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(policeoffice.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#0077B6", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Police Offices") +
  mapTheme()

E <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = shoppingcen, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(shoppingcen.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FF6662", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Shopping Centers") +
  mapTheme()


combined_maps <- A + B + C + D + E
#combined_maps <- combined_maps / plot_layout(nrow = 2, ncol = 3)
print(combined_maps)
```

## Census Data Trend

```{r Trends by Year, warning=FALSE, fig.width=8, fig.height=8}
a <- ggplot(allTracts, aes(x = as.factor(year), y = TotalPop, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Total Population: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase)

b <- ggplot(allTracts, aes(x = as.factor(year), y = TotalWorker, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Worker: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase)

c <- ggplot(allTracts, aes(x = as.factor(year), y = Workfromhome, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Work from Home: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase)
d <- ggplot(allTracts, aes(x = as.factor(year), y = Workwithbus, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of People Using Bus as Transportation to Work: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase)

combined_plot <- grid.arrange(a,b,c,d, ncol = 4, top = c(2, 2))

print(combined_plot)

```

### Commute by Bus Trend

```{r}
allTracts$year <- as.numeric(allTracts$year)

average_pct_bus <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bus = mean(pctworkwithbus))

ggplot(average_pct_bus, aes(x = year, y = avg_pct_bus)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Bus",
       title = "Trend of Commuting by Bus from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_car <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_car = mean(pctworkbycar))

ggplot(average_pct_car, aes(x = year, y = avg_pct_car)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Car",
       title = "Trend of Commuting by Car from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_taxi <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_taxi = mean(pctworkbytaxi))

ggplot(average_pct_taxi, aes(x = year, y = avg_pct_taxi)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Taxi",
       title = "Trend of Commuting by Taxi from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_motor <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_motor = mean(pctworkbymotor))

ggplot(average_pct_motor, aes(x = year, y = avg_pct_motor)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Motorcycle",
       title = "Trend of Commuting by Motorcycle from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_bike <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bike = mean(pctworkbybike))

ggplot(average_pct_bike, aes(x = year, y = avg_pct_bike)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Bike",
       title = "Trend of Commuting by Bike from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_walk <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_walk = mean(pctworkbywalk))

ggplot(average_pct_walk, aes(x = year, y = avg_pct_walk)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Walk",
       title = "Trend of Commuting by Walk from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

```


### Spatial Reference of Car Ownership and Commute by Bus

```{r, echo=FALSE, fig.height=18, fig.width=12}
tp_vars <- c("pctworkwithbus", "pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
tp_titles <- c("Percent of Population Commute with Bus", "Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2017,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2017 (Census Tracts)"))

# plot 2018
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2018,aes(fill = (!!sym(i))), crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2018 (Census Tracts)"))

# plot 2019
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2019,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2019 (Census Tracts)"))

#plot 2020
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2020,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2020 (Census Tracts)"))

#plot 2021
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2021,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2021 (Census Tracts)"))

#plot 2022
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2022,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "black", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j)))
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2022 (Census Tracts)"))
```

### Histogram of Car Ownership

```{r, echo=FALSE, fig.height=18, fig.width=12}
car_vars <- c("pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
car_titles <- c("Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
# hist 2017
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2017"))

# hist 2018
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2018, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j<-j+1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2018"))

# hist 2019
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2019, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j<-j+1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2019"))

# hist 2020
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2020, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j<-j+1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2020"))

# hist 2021
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2021, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j<-j+1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2021"))

# hist 2022
histList <- list()
j <- 1

for(i in car_vars){
  hist <- ggplot(data = dat2022, aes(x = !!sym(i))) +
           geom_histogram(fill = "#005357", color = "black") +
           labs(title = car_titles[j], caption = paste0("Figure ", as.character(j)))
  histList[[i]] <- hist
  j<-j+1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2022"))
```

### Car Ownership Trend

```{r}
average_pct_cars <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_0cars = mean(pctnocar),
            avg_pct_1cars = mean(pct1car),
            avg_pct_2cars = mean(pct2car),
            avg_pct_3cars = mean(pct3car),
            avg_pct_4cars = mean(pct4car),
            avg_pct_5ormore_cars = mean(pct5car))

average_pct_cars_long <- pivot_longer(average_pct_cars, 
                                      cols = starts_with("avg_pct"), 
                                      names_to = "Cars", 
                                      values_to = "Percentage")

plot_list <- list()
for (i in unique(average_pct_cars_long$Cars)) {
  num_cars <- gsub("avg_pct_", "", i)
  num_cars <- gsub("cars", "car", num_cars)
  subset_data <- subset(average_pct_cars_long, Cars == i)
  plot <- ggplot(subset_data, aes(x = year, y = Percentage, group = 1)) +
    geom_line(color = "black",linewidth=1.5) +
    geom_point(color = "red",size=3) +
    labs(x = "Year", 
         y = "Average Percentage of Households",
         title = paste("Trend of Car Ownership:", num_cars)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_minimal()
  plot_list[[i]] <- plot
}

grid.arrange(grobs = plot_list, ncol = 2, top = "Trend of Car Ownership (2017-2022)")
```



