---
title: "Combined Markdown"
author: "Teresa Chang, Roshini Ganesh, Jingyi Li, Jonathan Manurung"
date: "2024-02-27"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

### 1. Introduction

In 2021, the Charlotte Area Transit System (CATS) embarked on 'The Bus Priority Study' as part of the city's larger Strategic Mobility Plan, aiming to enhance bus trips' speed, reliability, and convenience for riders. To achieve this, CATS set out to understand recent dynamics, explore accessibility, and examine network efficiencies.

#### 1.1 CATS identified three primary objectives:

1. Understanding recent bus ridership dynamics involved 
2. Exploring thepresent reality of bus accessibility, considering ridership trends and demographic distributions.
3. Examining network efficiencies and possibilities 

#### 1.2 Actions required related to objectives
Identify underperforming bus stops
Study demographic, spatial, and economic factors influencing ridership and performance  
Alter current network of stops and routes  to optimize transit services.


### 2. Use case
Our team will support CATS transportation planners in improving bus line efficiency by analyzing historical ridership trends per bus stop to identify and predict underperforming stops and routes. 

With this, we will propose alternate network options through a web-based dashboard that presents predictions with the context of demographic, spatial, financial factors of influence. 


```{r setup, include=FALSE}
# You can set some global options for knitting chunks

knitr::opts_chunk$set(echo = TRUE)

# Load some libraries

library(tidyverse)
library(sf)
library(spdep)
library(caret)
library(ckanr)
library(FNN)
library(grid)
library(gridExtra)
library(ggcorrplot) # plot correlation plot
library(corrr)      # another way to plot correlation plot
library(kableExtra)
library(jtools)     # for regression model plots
library(ggstance) # to support jtools plots
library(ggpubr)    # plotting R^2 value on ggplot point scatter
library(broom.mixed) # needed for effects plots
library(tidycensus)
library(kableExtra)
library(stargazer)
library(readxl)
library(zoo)
library(cowplot)
library(patchwork)
library(randomForest)
library(caret)

# functions and data directory

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c")
palette7 <- c("white","#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c", "black")
red_palette <- c("#FFD3D3", "#FFA8A8", "#FF7D7D", "#FF5252", "#FF0000")
positive_palette <- c("darkgreen", "green", "yellow", "#FFAB00", "red")
negative_palette <- c("blue", "lightblue", "grey80", "grey60", "grey40")

positive_paletteaccent <- c("#006d2c", "#2ca25f", "#FFCB62", "#FFAB00", "#C60500")
negative_paletteaccent <- c("#0077B6", "#7ACFD3", "grey80", "grey60", "grey40")

palettebase5 <- c("#edf8fb","#7ACFD3","#3AA8AE","#2D8388","#005357")
palettebase7 <- c("#edf8fb","#7ACFD3","#54C1C6","#3AA8AE","#2D8388","#016F74","#005357")
palettebase <- c("#7ACFD3","#67C8C8","#54C1C6","#41BAC0","#3AA8AE","#33969B","#2D8388","#016F74","#005357")
paletteaccent <- c("#46AFB4","#249DA2","#01939A","#017278","#005A5E","#FFCB62","#FFBD37","#FFAB00","#C68500","#9C6900","#FF6662","#FF3C37","#FF0700","#C60500","#9C0400")

red_paletteaccent <- c("#FFD3D3","#FFA8A8","#FF0700","#C60500","#9C0400")
yellow_paletteaccent <- c("#FFE6A7","#FFBD37","#FFAB00","#C68500","#9C6900")

census_api_key("ab9309f9cc70c0e1895e7166c3ca981c40cf0331", overwrite = TRUE , install = TRUE)
```

### 3. Data 

#### 3.1 Datasets considered

To analyze and forecast performance trends, we have used ridership data from CATS for the years between 2017 and 2023. Additionally we have identified potential demographic, spatial, and economic parameters that may influence our predictor variable that is ridership. 

#### 3.2 Limitations of dataset

On initial review, certain limitations of the data present themselves. These include:
1. No hourly ridership data is available precluding us from analyzing peak hour trends
2. Weekend vs weekday ridership is not available which leads to the assumption of similar usage patterns of weekends, weekdays, and holidays
3. Data on stops added or removed in these years of study is not explicit 
4. Only stop locations are made available which means route locations will have to be sourced and spatially joined. This allows us to join only the most recent route data for these stops.
5. The dataset for 2023 is incomplete and only includes data for months January to November.


```{r cars, warning=FALSE, message=FALSE, include=FALSE}
stops<-read_csv('AllDat.csv')
bus_routes <- st_read("Data/Bus_Routes.geojson")%>%
  st_transform('EPSG:3358')
school <- st_read("Data/Schools.geojson")%>%
  st_transform('EPSG:3358')

parks <- st_read("Data/Parks.geojson")%>%
  st_transform('EPSG:3358')

groceries <-
  st_read("Data/Grocery_Stores.geojson")%>%
  st_transform('EPSG:3358')

shoppingcen <-
  st_read("Data/Existing_Shopping_Centers.geojson")%>%
  st_transform('EPSG:3358')

policeoffice <-
  st_read("Data/CMPD_Police_Division_Offices.geojson") %>%
  st_transform('EPSG:3358')

# Create a vector of years
years <- c(2017, 2018, 2019, 2020, 2021, 2022)

# Loop through the years to load variables
variables_list <- lapply(years, function(year) {
  load_variables(year, "acs5", cache = TRUE)
})

# Combine the variables from different years into a single data frame
all_variables <- do.call(rbind, variables_list)

hlt <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.geojson")

# 2017 census
dat2017 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2017, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2018
dat2018 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2018, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2018") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

#census 2019
dat2019 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2019, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2019") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2020
dat2020 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2020, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale)

# census 2021
dat2021 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2021, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2021") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2022
dat2022 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2022, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2022") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```

#### 3.3 Data Wrangling

```{r pressure, , warning=FALSE, message=FALSE, include=FALSE, echo=FALSE}
#feature engineering
school.sf <-
  school %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

parks.sf <-
  parks %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

groceries.sf <-
  groceries %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

shoppingcen.sf <-
  shoppingcen %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

policeoffice.sf <-
  policeoffice %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

silver_stops <- st_read("Data/LYNX_Silver_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

silver_routes <- st_read("Data/LYNX_Silver_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')

blue_stops <-
  st_read("Data/LYNX_Blue_Line_Stations.geojson")%>%
  st_transform('EPSG:3358')

blue_routes <-
  st_read("Data/LYNX_Blue_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

gold_stops <-
  st_read("Data/LYNX_Gold_Line_Stops.geojson")%>%
  st_transform('EPSG:3358')

gold_routes <-
  st_read("Data/LYNX_Gold_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

red_stops <-
  st_read("Data/LYNX_Red_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

red_routes <-
  st_read("Data/LYNX_Red_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')


allTracts <- rbind(
  dat2017,
  dat2018,
  dat2019,
  dat2020,
  dat2021,
  dat2022)
sf_allTracts <- st_as_sf(allTracts, wkt = "geometry")

bus_data <- stops %>% mutate(`Stop ID` = sprintf("%05s", `Stop ID`))
busstop_sf <- st_as_sf(bus_data, coords = c("Longitude", "Latitude"), crs = 4326)
catsstop_sf <- st_as_sf(stops, coords = c("Longitude", "Latitude"), crs = 2264)  # Assuming NAD 83/North Carolina (ftUS)
#Reproject catsstop_sf to WGS 84 (EPSG:4326)
catsstop_sf <- st_transform(catsstop_sf, 4326)
busstop_sf <- st_transform(busstop_sf, 3358)
catsstop_sf <- st_transform(catsstop_sf, 3358)
routes_sf <- st_transform(bus_routes, 3358)
allTracts_sf <- st_transform(allTracts, 3358)

bus_stops.sf <-
  stops %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = "EPSG:3358")%>%
    distinct()

```


#### 3.4 APC data

```{r, warning = FALSE, message=FALSE}
ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Board), size = .3) +
  scale_color_gradient(low = "#FFAB00", high = "darkgreen") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Board Value")

ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Alight), size = .3) +
  scale_color_gradient(low = "#FFAB00", high = "darkgreen") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Alight Value")

```

### 4. Initial Trends

1. Ridership does not seem to vary too much over the years across all stops based on the given CATS dataset. However, demographic data from ACS 2022 suggests that there has been a steep decline in the percentage of people who commute to work by bus over the last 7 years. This may be attributed to more wealthy people working from home along with more economically disadvantaged people choosing to take the bus. 
2. The global trend suggests that bus ridership during the pandemic did not drop significantly. We suspect that the consistent ridership profile may be due to errors in recording the data for pandemic years. 

#### 4.1 Performance before and after the COVID-19 pandemic

To investigate the nature of performance before and after pandemic years further, the difference between average ridership is mapped pre Covid (2017- 2020/3) and post Covid (2020/3 - 2023). This mapping exercise indicates that ridership has decreased in most bus stops post the pandemic. Additionally, some stops show a hundred percent decline. These are filtered out because it may indicate that the stop is no longer in service.

Next, stops with an increase in ridership are separated from stops with a decrease and the results are mapped. This suggests that a majority of stops in the city center have experienced a decrease in ridership. This spatial clustering of underperforming stops suggests areas for targeted intervention.


```{r, warning=FALSE, message=FALSE, include=FALSE}
# threshold: date of covid bgan
threshold_date <- as.Date("2020-03-01") 

stops_data <- stops %>%
  mutate(Numeric_Month = match(Month, month.name)) %>%
  mutate(Readable_Date = as.yearmon(paste(Year, sprintf("%02d", Numeric_Month)), "%Y %m")) %>%
  mutate(Covid_Status = ifelse(as.Date(Readable_Date) >= threshold_date, "Post", "Pre"))%>%
  rename(Stop_Name = `Stop Name`)%>%
  rename(Stop_ID = `Stop ID`)

# Calculate the increase/decrease
average_board <- stops_data %>%
  group_by(Stop_Name, Numeric_Month, Covid_Status) %>%
  summarize(Avg_Board = mean(Avg_Board, na.rm = TRUE))

average_board_pivoted <- pivot_wider(average_board, names_from = Covid_Status, values_from = Avg_Board)

average_board_pivoted <- average_board_pivoted %>%
  mutate(Difference = (Post - Pre)/Pre *100 ) %>%
  select(Stop_Name, Numeric_Month, Pre, Post, Difference)

average_board_pivoted <- na.omit(average_board_pivoted)
# join the longitude and latitude data back and select the first one
average_board_pivoted <- average_board_pivoted %>%
  left_join(stops_data %>% select(Stop_Name, Longitude, Latitude) %>% distinct(Stop_Name, .keep_all = TRUE), by = "Stop_Name")

# Adjust the month as needed
selected_month <- 1  
selected_data <- average_board_pivoted %>%
  filter(Numeric_Month == selected_month)

stops_sf <- st_as_sf(selected_data, coords = c("Longitude", "Latitude"), crs = 4326)
stops_sf <- st_transform(stops_sf, crs = 3358)  # NAD83 North Carolina

```

##### 4.1.1 All stops from 2017\~2023

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
palettebase5_rev <- rev(palettebase5)

#ggplot() +
#  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
#  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
#  geom_sf(data = stops_sf , aes(colour = q5(Difference)), 
#          show.legend = "point", size = 1, alpha= 0.5) +
#  scale_colour_manual(values = palettebase5_rev,
#                   labels=qBr(stops_sf ,"Difference"),
#                   name="Quintile\nBreaks (%)") +
#  labs(title = paste("Difference (%) - Month:", selected_month),
#       fill = "Difference (Post - Pre)") +
#  mapTheme()

filtered_stops_sf <- stops_sf[stops_sf$Difference != -100, ]

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = filtered_stops_sf , aes(colour = q5(Difference)), 
          show.legend = "point", size = 1, alpha= 0.5) +
  scale_colour_manual(values = palettebase5_rev,
                   labels=qBr(filtered_stops_sf ,"Difference"),
                   name="Quintile\nBreaks (%)") +
  labs(title = paste("Difference (%) (-100% excluded)   - Month:", selected_month),
       fill = "Difference (Post - Pre)") +
  mapTheme()
```

##### 4.1.2 Increase vs Decrease

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
stops_sf$PositiveDifference <- stops_sf$Difference > 0
positive_stops_sf <- stops_sf[stops_sf$PositiveDifference,]
negative_stops_sf <- stops_sf[!stops_sf$PositiveDifference,]

yellow_paletteaccent_rev <- rev(yellow_paletteaccent)

# positive and negative differences
positive_plot <- ggplot(positive_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = red_paletteaccent,
                     labels = qBr(positive_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Positive Increase (%) - Month:", selected_month),
       fill = "Positive Difference (Post - Pre)") +
  mapTheme()

negative_plot <- ggplot(negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()

#combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)
#print(combined_plot)

# Exclude -100%
filtered_negative_stops_sf <- negative_stops_sf[negative_stops_sf$Difference != -100, ]

negative_plot <- ggplot(filtered_negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(filtered_negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) (-100% excluded) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()
combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)

print(combined_plot)
```

> From the plotted maps, there is a majority of stops in Charlotte’s city center that have experienced a decrease in ridership before and after the COVID-19 pandemic



#### 4.2 Top 50 underperforming stops from 2017 to 2023

Between 2017-2023, there is a uniform trend of underperformance in certain underperforming stops. Other underperforming stops exhibit a growing trend of underperformance post the pandemic. About half of these stops exhibit spatial clustering patterns and are located along a single route.


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
# filter underperforming stops
underperforming_stops <- stops_data[(stops_data$Board + stops_data$Alight) <= 10, ]
underperformance_counts <- table(underperforming_stops$Stop_ID)


top_50_stops <- names(sort(underperformance_counts, decreasing = TRUE)[1:50])
underperforming_top_50 <- underperforming_stops[underperforming_stops$Stop_ID %in% top_50_stops, ]
underperforming_top_50 <- underperforming_top_50 %>%
  mutate(Total_Board_Alight = Board + Alight)

# Order
underperforming_top_50$Stop_ID <- factor(underperforming_top_50$Stop_ID, levels = rev(top_50_stops))

stacked_bar_plot <- ggplot(underperforming_top_50, aes(x = Stop_ID, fill = factor(Year))) +
  geom_bar(stat = "count", position = "stack", color = "darkgrey" ) +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = palettebase7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels
## print(stacked_bar_plot) 

underperforming_top_50_all_sf <- st_as_sf(underperforming_top_50, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_all_sf <- st_transform(underperforming_top_50_all_sf, crs = 3358)  # NAD83 North Carolina


map_plot_top_50 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_all_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_all_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts (2017 - 2023)",
       fill = "Avg_Board + Avg_Alight") +
  mapTheme()
## print(map_plot_top_50) 

combined_plot <- stacked_bar_plot + map_plot_top_50
print(combined_plot)
```

#### 4.3 Top 50 underperforming stops in 2022

The underperformance and spatial clustering trends in 2023 are consistent with the overall trend.

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
# Underperforming Stops in 2022
underperforming_stops_2022 <- stops_data[(stops_data$Board + stops_data$Alight) <= 10 & stops_data$Year == 2022, ]
underperformance_counts_2022 <- table(underperforming_stops_2022$Stop_ID)

# Top 30 Stops for 2023
top_50_stops_2022 <- names(sort(underperformance_counts_2022, decreasing = TRUE)[1:50])
underperforming_top_50_2022 <- underperforming_stops_2022[underperforming_stops_2022$Stop_ID %in% top_50_stops_2022, ]


# Order
underperforming_top_50_2022$Stop_ID <- factor(underperforming_top_50_2022$Stop_ID, levels = rev(top_50_stops_2022))

bar_plot_top_50_2022 <- ggplot(underperforming_top_50_2022, aes(x = Stop_ID, fill = "2022")) +
  geom_bar(stat = "count", color = "black") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = c("2022" = "#7ACFD3"), name = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
##print(bar_plot_top_50_2022)

underperforming_top_50_2022 <- underperforming_top_50_2022 %>%
  mutate(Total_Board_Alight = Board + Alight)


underperforming_top_50_sf <- st_as_sf(underperforming_top_50_2022, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_sf <- st_transform(underperforming_top_50_sf, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_2022 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts in 2022",
       fill = "Total_Board_Alight") +
  mapTheme()
##print(map_plot_top_50_2022)

combined_plot <- bar_plot_top_50_2022 + map_plot_top_50_2022
print(combined_plot)
```


> Some stops exhibit a growing trend of underperformance post the pandemic.About half of these stops exhibit spatial clustering patterns and are located along a single route. The underperformance and clustering trends in 2022 are consistent with the overall trend in the city center area.


```{r, warning=FALSE, message=FALSE, fig.height=10, fig.width=8}
##print(map_plot_top_50) 
##print(map_plot_top_50_2022)

##print(stacked_bar_plot) 
##print(bar_plot_top_50_2022)
```

### 5. Spatial factors:

#### 5.1 Existing and proposed transit lines and stations

Other existing and proposed transit stops and routes are considered to study the relationship between infrastructure and underperforming stops, if any. If significant, the distance of bus stops to other transit stops would also serve as variables in the model.


```{r, warning=FALSE, message=FALSE}
transit <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, aes(color = "Bus Route"), size = 5, alpha = 0.5) +
  geom_sf(data = blue_routes, aes(color = "Blue Line"), size = 8) +
  geom_sf(data = blue_stops, aes(color = "Blue Line"), size = 2) +
  geom_sf(data = gold_routes, aes(color = "Gold Line"), size = 8) +
  geom_sf(data = gold_stops, aes(color = "Gold Line"), size = 2) +
  geom_sf(data = silver_routes, aes(color = "Silver Line"), size = 8, linetype = "dashed") +
  geom_sf(data = silver_stops, aes(color = "Silver Line"), size = 2, shape = 21) +
  geom_sf(data = red_routes, aes(color = "Red Line"), size = 8, linetype = "dashed") +
  geom_sf(data = red_stops, aes(color = "Red Line"), size = 2, shape = 21) +
  
  labs(title = "Existing and Proposed Transit Lines and Stops") +
    scale_color_manual(name = "Lines",
                     values = c("Bus Route" = "darkgrey",
                                "Blue Line" = "#4682B4", 
                                "Gold Line" = "#FFD700", 
                                "Silver Line" = "purple", 
                                "Red Line" = "#FF6347"),
                     labels = c("Blue Line","Bus Routes", "Gold Line", "Proposed Red Line", "Proposed Silver Line")) +

  mapTheme()

print(transit)

```

#### 5.2 Density of Amenities

Most amenities are clustered within the city center and expanded to the south, presenting an inequitable distribution. For model input, the distance to the amenities would serve as variables as well

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
A <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = school, color = "darkgrey", size = .5, alpha = 0.3) +
  stat_density2d(data = data.frame(st_coordinates(school.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFE6A7", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Schools") +
  mapTheme()

B <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = parks, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(parks.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#7ACFD3", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Parks") +
  mapTheme()

C <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = groceries, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(groceries.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFAB00", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Grocery Stores") +
  mapTheme()

D <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = policeoffice, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(policeoffice.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#0077B6", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Police Offices") +
  mapTheme()

E <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = shoppingcen, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(shoppingcen.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FF6662", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Shopping Centers") +
  mapTheme()


combined_maps <- A + B + C + D + E
#combined_maps <- combined_maps / plot_layout(nrow = 2, ncol = 3)
print(combined_maps)
```

### 6. Demographic Factors

#### 6.1 Mode Trends 

Almost all census tracts with a higher percentage of population that commute by bus to work are located close to bus stops. However, some suburban census tracts with a high percentage of population that commute by bus are far from bus stops. This may be considered as a variable that influences prediction.

##### 6.1.1 By Year and Population

```{r Trends by Year, warning=FALSE, message=FALSE, fig.width=8, fig.height=4}
a <- ggplot(allTracts, aes(x = as.factor(year), y = TotalPop, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Total Population: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

b <- ggplot(allTracts, aes(x = as.factor(year), y = TotalWorker, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Worker: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

c <- ggplot(allTracts, aes(x = as.factor(year), y = Workfromhome, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Work from Home: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)
d <- ggplot(allTracts, aes(x = as.factor(year), y = Workwithbus, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of People Using Bus as Transportation to Work: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

combined_plot <- grid.arrange(a,b,c,d, ncol = 4, top = c(2, 2))

print(combined_plot)

```

##### 6.1.2 Mode Share

The general trend of percentage of population commute by bus has decreased. The 2022 percentage is halved comparing to 2017 percentage. In comparison, the general trend of percentage of population commute by car has decreased as by more than 10%. Additionally, the general trend of percentage of population commute by bicycle has a subtle decrease. Over the 6 year, it decreased 0.07%. Finally, the general trend of percentage of population commute by walk has slightly increased. Its peak was during Covid-19, but post Covid-19 period, the rate drops.

```{r, warning=FALSE, message=FALSE}
allTracts$year <- as.numeric(allTracts$year)

average_pct_bus <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bus = mean(pctworkwithbus))

ggplot(average_pct_bus, aes(x = year, y = avg_pct_bus)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Bus",
       title = "Trend of Commuting by Bus from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_car <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_car = mean(pctworkbycar))

ggplot(average_pct_car, aes(x = year, y = avg_pct_car)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Car",
       title = "Trend of Commuting by Car from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_taxi <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_taxi = mean(pctworkbytaxi))

ggplot(average_pct_taxi, aes(x = year, y = avg_pct_taxi)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Taxi",
       title = "Trend of Commuting by Taxi from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_motor <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_motor = mean(pctworkbymotor))

ggplot(average_pct_motor, aes(x = year, y = avg_pct_motor)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Motorcycle",
       title = "Trend of Commuting by Motorcycle from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_bike <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bike = mean(pctworkbybike))

ggplot(average_pct_bike, aes(x = year, y = avg_pct_bike)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Bike",
       title = "Trend of Commuting by Bike from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_walk <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_walk = mean(pctworkbywalk))

ggplot(average_pct_walk, aes(x = year, y = avg_pct_walk)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Walk",
       title = "Trend of Commuting by Walk from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

```


##### 6.1.3 Demographic Maps

The majority of Mecklenburg County population resides in suburban census tract, a trend that is consistent with the number of employed individuals in the county who also live in the suburbs. However, when examining census data on the use of public transportation to commute to work, a significant proportion is concentrated in the city center and expanded to the south. This can be correlated with census data on the prevalence of working from home, which tends to be dispersed in suburban areas situated relatively farther away from the city center


```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12}
tp_vars2 <- c("TotalPop", "TotalWorker", "Workwithbus", "Workfromhome" )
tp_titles2 <- c("Total Population ", "Total Population of Worker", "Total Population Commute with Bus", "Total Population Work from Home ")
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2017,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2017 (Census Tracts)"))

# plot 2018
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2018,aes(fill = (!!sym(i))), crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2018 (Census Tracts)"))

# plot 2019
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2019,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2019 (Census Tracts)"))

#plot 2020
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2020,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2020 (Census Tracts)"))

#plot 2021
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2021,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2021 (Census Tracts)"))

#plot 2022
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2022,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2022 (Census Tracts)"))
```

> From the plotted maps, suburban census tracts in Charlotte have more residents as well as more number of employed workers. For the people commuting to work by bus are largely concentrated in the city center of Charlotte. Meanwhile, work from home trends indicate dominant concentrations in the suburban areas.


##### 6.1.4 Spatial Reference of Car Ownership and Commute by Bus

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12}
tp_vars <- c("pctworkwithbus", "pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
tp_titles <- c("Percent of Population Commute with Bus", "Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2017,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2017 (Census Tracts)"))

# plot 2018
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2018,aes(fill = (!!sym(i))), crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2018 (Census Tracts)"))

# plot 2019
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2019,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2019 (Census Tracts)"))

#plot 2020
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2020,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2020 (Census Tracts)"))

#plot 2021
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2021,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2021 (Census Tracts)"))

#plot 2022
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2022,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2022 (Census Tracts)"))
```


##### 6.1.5 Histogram of Car Ownership

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12}
car_vars <- c("pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
car_titles <- c("Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
# hist 2017
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2017"))

# hist 2018
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2018"))

# hist 2019
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2019"))

# hist 2020
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2020"))

# hist 2021
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2021"))

# hist 2022
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2022"))
```

> From the plotted maps, suburban census tracts in Charlotte have more households with cars than household in city center. Most of those households have two or more cars


##### 6.1.6 Car Ownership Trend

```{r, warning=FALSE, message=FALSE}
average_pct_cars <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_0cars = mean(pctnocar),
            avg_pct_1cars = mean(pct1car),
            avg_pct_2cars = mean(pct2car),
            avg_pct_3cars = mean(pct3car),
            avg_pct_4cars = mean(pct4car),
            avg_pct_5ormore_cars = mean(pct5car))

average_pct_cars_long <- pivot_longer(average_pct_cars, 
                                      cols = starts_with("avg_pct"), 
                                      names_to = "Cars", 
                                      values_to = "Percentage")

plot_list <- list()
for (i in unique(average_pct_cars_long$Cars)) {
  num_cars <- gsub("avg_pct_", "", i)
  num_cars <- gsub("cars", "car", num_cars)
  subset_data <- subset(average_pct_cars_long, Cars == i)
  plot <- ggplot(subset_data, aes(x = year, y = Percentage, group = 1)) +
    geom_line(color = "black",linewidth=1.5) +
    geom_point(color = "red",size=3) +
    labs(x = "Year", 
         y = "Average Percentage of Households",
         title = paste("Trend of Car Ownership:", num_cars)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_minimal()
  plot_list[[i]] <- plot
}

grid.arrange(grobs = plot_list, ncol = 2, top = "Trend of Car Ownership (2017-2022)")
```

### 7. Next Steps

1.To identify most significant demographic and spatial variables for predicting underperformance
2. To present findings to clients and solicit feedback
3. To arrive at best modeling process to predict stops that are likely to underperform
4. To propose removal of and addition of stops to certain routes and examine their equity and financial impact
5. To create a web-based dashboard that allows CATS transportation planners to overlay socioeconomic, equity, and financial factors with underperformance to be able to allocate resources to stops effectively

In doing so, our aim is to build a comprehensive and robust model to optimize bus line efficiency, aligning with CATS' goal of providing faster, more reliable, and convenient transit options for Charlotte residents.


```{r, warning=FALSE, message=FALSE}
# Perform spatial join based on containment (bus stops within census tracts)
joined_data <- st_join(busstop_sf, allTracts_sf, join = st_within)
```


```{r, warning=FALSE, message=FALSE}
a# Spatial join based on containment (bus stops within census tracts)
if (st_crs(busstop_sf) != st_crs(allTracts_sf)) {
  busstop_sf <- st_transform(busstop_sf, crs = st_crs(allTracts_sf))
}

# Spatial join
joined_data <- st_join(busstop_sf, allTracts_sf, join = st_within)

# Separate dataset
data_2017 <- filter(joined_data, Year == 2017 & year == 2017)
data_2018 <- filter(joined_data, Year == 2018 & year == 2018)
data_2019 <- filter(joined_data, Year == 2019 & year == 2019)
data_2020 <- filter(joined_data, Year == 2020 & year == 2020)
data_2021 <- filter(joined_data, Year == 2021 & year == 2021)
data_2022 <- filter(joined_data, Year == 2022 & year == 2022)

# Save csv
#write.csv(data_2017, "data_2017.csv", row.names = FALSE)
#write.csv(data_2018, "data_2018.csv", row.names = FALSE)
#write.csv(data_2019, "data_2019.csv", row.names = FALSE)
#write.csv(data_2020, "data_2020.csv", row.names = FALSE)
#write.csv(data_2021, "data_2021.csv", row.names = FALSE)
#write.csv(data_2022, "data_2022.csv", row.names = FALSE)
```


```{r clean_school}
#joined_data$school.Buffer <- joined_data %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(school.sf, counter = 1),., sum) %>%
#    pull(counter)

#joined_data$policeoffice.Buffer <- joined_data %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(policeoffice.sf, counter = 1),., sum) %>%
#    pull(counter)

#data_2022$busstop.Buffer <- data_2022 %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(bus_stops.sf, counter = 1),., sum) %>%
#    pull(counter)


data_2022 <-
  data_2022 %>% 
    mutate(
      school_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 1),
      
      school_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 2), 
      
      school_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 3), 
      
      school_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 4), 
      
      school_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      policeoffice_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 1),
      
      policeoffice_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 2), 
      
      policeoffice_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 3), 
      
      policeoffice_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 4), 
      
      policeoffice_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      groceries_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 1),
      
      groceries_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 2), 
      
      groceries_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 3), 
      
      groceries_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 4), 
      
      groceries_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 5)) 


data_2022 <-
  data_2022 %>% 
    mutate(
      parks_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 1),
      
      parks_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 2), 
      
      parks_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 3), 
      
      parks_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 4), 
      
      parks_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      shoppingcen_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 1),
      
      shoppingcen_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 2), 
      
      shoppingcen_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 3), 
      
      shoppingcen_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 4), 
      
      shoppingcen_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 5)) 

```

```{r}
# Rename OBJECTID_1 column to OBJECTID in blue_stops
blue_stops <- blue_stops %>% rename(OBJECTID = OBJECTID_1)
blue_stops <- select(blue_stops, c(OBJECTID, NAME))

gold_stops <- select(gold_stops, c(OBJECTID, Stop_Name))
gold_stops <- gold_stops %>% rename(NAME = Stop_Name)
# Merge the datasets based on OBJECTID
existing_transit_stops <- rbind(blue_stops, gold_stops)
existing_transit_stops <- st_as_sf(existing_transit_stops, coords = c("geometry.x", "geometry.y"))


# Plot the density of points
ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = existing_transit_stops, color = "black", size = .5, alpha = 0.5) +
  stat_density_2d(data = existing_transit_stops, 
                  aes(x = st_coordinates(geometry)[, 1], 
                      y = st_coordinates(geometry)[, 2],
                      fill = ..level.., alpha = ..level..),
                  size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "#737373", high = "#0077B6", name = "Density") +  # Adjust color gradient
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Existing Light Rail Stops") +
  mapTheme()

```

```{r}
data_2022 <-
  data_2022 %>% 
    mutate(
      transit_stops_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 1),
      
      transit_stops_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 2), 
      
      transit_stops_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 3), 
      
      transit_stops_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 4), 
      
      transit_stops_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 5)) 
```

```{r}
city_center<- st_sfc(st_point(c(1692115, 5236081)), crs = 3358)
data_2022$dist_to_center <- st_distance(data_2022$geometry, city_center)
```


### Modeling

#### OLS

```{r}
data_2022 <- data_2022 %>% 
             mutate(Avg_ridership = Avg_Board + Avg_Alight)
# split traning and testing
set.seed(123) # Set seed for reproducibility
train_index <- createDataPartition(data_2022$Avg_ridership, p = 0.7, list = FALSE)
OLS_train_data <- data_2022[train_index, ]
OLS_test_data <- data_2022[-train_index, ]

#Fit regression
OLS <- lm(Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center
          , data = OLS_train_data)

# Evaluate model
OLS_testing <- OLS_test_data %>%
  mutate(
    Predictions = predict(OLS, OLS_test_data),
    Error = Predictions - OLS_test_data$Avg_ridership,
    AbsError = abs(Predictions - OLS_test_data$Avg_ridership),
    MAE = mean(abs(Error), na.rm = TRUE),
    APE = abs(Predictions - OLS_test_data$Avg_ridership) / Predictions)
summary(OLS)
```

```{r}
plot(OLS)
```

```{r}
# Extract actual and predicted values
OLSactual <- OLS_test_data$Avg_ridership
OLSpredicted <- predict(OLS, newdata = OLS_test_data)

# Create a scatter plot of actual vs. predicted values
plot(OLSactual, OLSpredicted, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


```{r}
OLS_testing_sf <- st_as_sf(OLS_testing)

ggplot(OLS_testing_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


#### Random Forest

```{r} 
# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index <- sample(1:nrow(data_2022), 0.7 * nrow(data_2022)) 
train_data_RF <- data_2022[train_index, ]
train_data_RF <- na.omit(train_data_RF)
test_data_RF <- data_2022[-train_index, ]
test_data_RF<-na.omit(test_data_RF)

# Fit the Random Forest model
RF_model <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5 + Month + PopDensity_km2, 
  data = train_data_RF)


RF_testing <- test_data_RF %>%
  mutate(Predictions_RF = predict(RF_model, newdata = test_data_RF),
Error_RF = Predictions_RF - test_data_RF$Avg_ridership,
AbsError_RF = abs(Predictions_RF - test_data_RF$Avg_ridership),
MAE_RF= mean(abs(Error_RF)),
APE_RF = (abs(Predictions_RF - test_data_RF$Avg_ridership)) / Predictions_RF)

RF_model
```

```{r}
# Extract actual and predicted values
actual <- test_data_RF$Avg_ridership
predicted <- predict(RF_model, newdata = test_data_RF)

# Create a scatter plot of actual vs. predicted values
plot(actual, predicted, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```

```{r}
RF_testing_sf <- st_as_sf(RF_testing)

ggplot(RF_testing_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


##### K-fold validation

> Random Forest 1

```{r}
# Load required packages
install.packages("caret")
library(caret)

# Define the training control parameters for cross-validation
ctrl <- trainControl(method = "cv", 
                     number = 10,     # Number of folds
                     verboseIter = TRUE)  

# Train the random forest model using k-fold cross-validation
RF_model <- train(Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5, 
                  data = train_data_RF,
                  method = "rf",  
                  trControl = ctrl)  

# Assess the performance of the model
summary(RF_model)
```

```{r}
RF_model$resample[1:5,]

mean(RF_model$resample[,3])
```


#### Random Forest#2

```{r}

routes_test <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.shp")

new_routes <- routes_test[, c("StopID", "routes")]


# Assuming your dataset is named "new_data" and column "a" contains values like "00040", "00125", etc.
# Convert column "a" to numeric and then format it without leading zeros
new_routes_2 <- new_routes %>%
  mutate(StopID = as.numeric(StopID),
         StopID = sprintf("%d", StopID))


new_routes_2 <- new_routes_2 %>%
  rename('Stop ID' = StopID)


# Convert the sf object "sf_data" to a regular data frame
routes_df <- st_drop_geometry(new_routes_2)

# Remove spaces in "Stop ID" values
data_2022$`Stop ID` <- gsub("\\s+", "", data_2022$`Stop ID`)

# Perform a left join based on the common column "Stop ID"
merged_dataset_bus <- left_join(routes_df, data_2022,   by = "Stop ID")

merged_dataset_bus <- merged_dataset_bus %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


# Separate the routes into individual rows
separated_data <- merged_dataset_bus %>%
  separate_rows(routes, sep = ",\\s*")

# Count the number of routes in each row
# merged_dataset_bus$total_of_routes <- sapply(strsplit(merged_dataset_bus$routes, ",\\s*"), length)


# Calculate the count of unique Stop ID values for each route
# count_stops_data <- separated_data %>% group_by(routes, `Stop ID`) %>% summarise(total_stops_in_routes = n_distinct(`Stop ID`))

count_stops_routes <- separated_data %>%
  group_by(routes) %>%
  summarise(total_stops_in_routes = n_distinct(`Stop ID`))

# Merge the count_stops_data back to the original_data based on routes and counts
merge_count_data <- left_join(separated_data, count_stops_routes, by = "routes")

merge_count_data <- merge_count_data %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


summarized_data2 <- merge_count_data %>%
  group_by(`Stop ID`, Month) %>%
  summarize(total_stops_in_served_routes = sum(total_stops_in_routes, na.rm = TRUE))


# Merge the summarized data back with the original 'bus_2022' dataset
RF2_table <- merge(merged_dataset_bus, summarized_data2, by = c('Stop ID', 'Month'), all.x = TRUE)


RF2_table_sf <- st_as_sf(RF2_table)


```


```{r} 

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index2 <- sample(1:nrow(RF2_table_sf), 0.7 * nrow(RF2_table_sf)) 
train_data_RF2 <- RF2_table_sf[train_index2, ]
train_data_RF2 <- na.omit(train_data_RF2)
test_data_RF2 <- RF2_table_sf[-train_index2, ]
test_data_RF2<- na.omit(test_data_RF2)


# Fit the Random Forest model
RF_model2 <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + Month + PopDensity_km2 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center, 
  data = train_data_RF2)


RF_testing2 <- test_data_RF2 %>%
  mutate(Predictions_RF2 = predict(RF_model2, newdata = test_data_RF2),
Error_RF2 = Predictions_RF2 - test_data_RF2$Avg_ridership,
AbsError_RF2 = abs(Predictions_RF2 - test_data_RF2$Avg_ridership),
MAE_RF2= mean(abs(Error_RF2)),
APE_RF2 = (abs(Predictions_RF2 - test_data_RF2$Avg_ridership)) / Predictions_RF2)

RF_model2
```


```{r}
# Extract actual and predicted values
actual2 <- test_data_RF2$Avg_ridership
predicted2 <- predict(RF_model2, newdata = test_data_RF2)

# Create a scatter plot of actual vs. predicted values
plot(actual2, predicted2, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
RF_testing_sf2 <- st_as_sf(RF_testing2)

ggplot(RF_testing_sf2) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF2), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
# Calculating Monthly Prediction
RF_testing_sf_monthly <- RF_testing_sf2 %>%
  mutate(Monthly_prediction = Predictions_RF2 * Number_of_Days)

file_path <- "predictions.geojson"
st_write(RF_testing_sf_monthly, file_path, driver = "GeoJSON")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
underperforming_stops_predicted2 <- RF_testing_sf_monthly[RF_testing_sf_monthly$Monthly_prediction <= 10, ]
underperformance_counts_predicted2 <- table(underperforming_stops_predicted2$"Stop ID")

# Top 50 Stops for 2023
top_50_stops_predicted2 <- names(sort(underperformance_counts_predicted2, decreasing = TRUE)[1:50])
underperforming_top_50_predicted2 <- underperforming_stops_predicted2[underperforming_stops_predicted2$"Stop ID" %in% top_50_stops_predicted2, ]

# Order
underperforming_top_50_predicted2$"Stop ID" <- factor(underperforming_top_50_predicted2$"Stop ID", levels = rev(top_50_stops_predicted2))

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, coords = c("Longitude", "Latitude"), crs = 4326)

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, wkt = "geometry", crs = 4326)

#underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)




underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_predicted2 <- ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_predicted2, aes(geometry = geometry, colour = q5(Monthly_prediction)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = c("1", "2", "3", "4", "5"),
                     name = "Quintile\nBreaks") +
  labs(title = "50 Bus Stops with Lowest Predicted Avg Ridership",
       fill = "Predictions_RF2") +
  mapTheme()+
  theme(legend.position = "bottom")

print(map_plot_top_50_predicted2)

file_path <- "Top50underperforming.geojson"
st_write(underperforming_top_50_predicted2, file_path, driver = "GeoJSON")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}
underperforming_stops_predicted2_m <- underperforming_stops_predicted2[underperforming_stops_predicted2$Month == 'January', ]

ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_stops_predicted2_m, aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January", color = "Predicted Ridership") +
  theme(legend.position = "bottom")

file_path <- "underperforming_Jan.geojson"
st_write(underperforming_stops_predicted2_m, file_path, driver = "GeoJSON")
```



```{r}
# Random Forrest 2 multicolinearity 

library(car)

# Get variable importance scores from the Random Forest model
importance_scores <- importance(RF_model2)

# Print the variable importance scores
print(importance_scores)

# Plot the variable importance scores
varImpPlot(RF_model2)
```


```{r}
# Load required packages
install.packages("caret")
library(caret)

# Define the training control parameters for cross-validation
ctrl2 <- trainControl(method = "cv", 
                     number = 5,     # Number of folds
                     verboseIter = TRUE)  

# Train the random forest model using k-fold cross-validation
RF_model2 <- train(Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + Month + PopDensity_km2 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center, 
                  data = train_data_RF2,
                  method = "rf",  
                  trControl = ctrl2)  

# Assess the performance of the model
summary(RF_model2)
```

```{r}
RF_model2$resample[1:5,]

mean(RF_model2$resample[,3])
```



#### Use-Case Scenario


##### Scenario 1: Increase in Population Density

###### Increase in Population Density 15% (2022 & 2017)
The last measured population density for Charlotte, NC was 2,827 in 2018. Charlotte, NC experienced an average growth rate of 2.87% from our first statistic recorded in 2009. If past trends continue, we forecast the population density growth rate would be 15% from 2022 to 2027.

> Random Forest #1

```{r, scenario 1 2022}

# Simulate 10% increase in PopDensity
data_2022_scenario1 <- data_2022 %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.15)

# Make Predictions with the New Scenario Data
predictions_2022_scenario1 <- predict(RF_model, newdata = data_2022_scenario1)

# Add Predictions to the Data
data_2022_scenario1$Predicted_Avg_ridership <- predictions_2022_scenario1

# Format Predicted Avg Ridership as Decimal Numbers
data_2022_scenario1$Predicted_Avg_ridership <- sprintf("%.2f", predictions_2022_scenario1)

# Convert Data to an sf Object
data_2022_scenario1_sf <- st_as_sf(data_2022_scenario1, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
data_2022_scenario1_sf$Predicted_Avg_ridership <- as.numeric(data_2022_scenario1_sf$Predicted_Avg_ridership)

# Plot the Map with Correct Fill Scale
ggplot(data_2022_scenario1_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Predicted_Avg_ridership), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops 2022", color = "Predicted Avg Ridership with \n15% Increase in Population Density")

```


> Random Forest #2

```{r, scenario 1 2022, warning=FALSE, message=FALSE, fig.height=8, fig.width=16}

# Simulate 15% increase in PopDensity
RF2_table_2022_scenario1 <- RF2_table_sf %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.15)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario1 <- predict(RF_model2, newdata = RF2_table_2022_scenario1)

# Add Predictions to the Data
RF2_table_2022_scenario1$Predicted_Avg_ridership <- predictions2_2022_scenario1

# Add Monthly Prediction Column
RF2_table_2022_scenario1 <- RF2_table_2022_scenario1 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
#RF2_table_2022_scenario1$Monthly_prediction <- sprintf("%.2f", predictions2_2022_scenario1)

# Convert Data to an sf Object
RF2_table_2022_scenario1_sf <- st_as_sf(RF2_table_2022_scenario1, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario1_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario1_sf$Monthly_prediction)

result <- RF2_table_2022_scenario1_sf[RF2_table_2022_scenario1_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
#  geom_sf(data = dat2022,aes(fill = (PopDensity_km2)), alpha = 0.5) +
#  scale_fill_gradient(low = "white", high = "grey20") + 
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario I", color = "Predicted Avg Ridership with \n15% Increase in Population Density")+
  theme(legend.position = "bottom")


#file_path <- "scenario1underperforming.geojson"
#st_write(result, file_path, driver = "GeoJSON")



```


###### Increase in Population Density 30%

> Random Forest #1

```{r, scenario 1a 2022}

# Simulate 30% increase in PopDensity
data_2022_scenario1a <- data_2022 %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.3)

# Make Predictions with the New Scenario Data
predictions_2022_scenario1a <- predict(RF_model, newdata = data_2022_scenario1a)

# Add Predictions to the Data
data_2022_scenario1a$Predicted_Avg_ridership <- predictions_2022_scenario1a

# Format Predicted Avg Ridership as Decimal Numbers
data_2022_scenario1a$Predicted_Avg_ridership <- sprintf("%.2f", predictions_2022_scenario1a)

# Convert Data to an sf Object
data_2022_scenario1a_sf <- st_as_sf(data_2022_scenario1a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
data_2022_scenario1a_sf$Predicted_Avg_ridership <- as.numeric(data_2022_scenario1a_sf$Predicted_Avg_ridership)

# Plot the Map with Correct Fill Scale
ggplot(data_2022_scenario1a_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Predicted_Avg_ridership), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops 2022", color = "Predicted Avg Ridership with \n30% Increase in Population Density")


```


> Random Forest #2


```{r, scenario 1a 2022}

# Simulate 30% increase in PopDensity
RF2_table_2022_scenario1a <- RF2_table_sf %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.3)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario1a <- predict(RF_model2, newdata = RF2_table_2022_scenario1a)

# Add Predictions to the Data
RF2_table_2022_scenario1a$Predicted_Avg_ridership <- predictions2_2022_scenario1a

# Add Monthly Prediction Column
RF2_table_2022_scenario1a <- RF2_table_2022_scenario1a %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario1a$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario1a)

# Convert Data to an sf Object
RF2_table_2022_scenario1a_sf <- st_as_sf(RF2_table_2022_scenario1a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario1a_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario1a_sf$Monthly_prediction)


result <- RF2_table_2022_scenario1a_sf[RF2_table_2022_scenario1a_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario I", color = "Predicted Avg Ridership with \n30% Increase in Population Density")+
  theme(legend.position = "bottom")

#file_path <- "scenario1aunderperforming.geojson"
#st_write(result, file_path, driver = "GeoJSON")


```


##### Scenario 2: Increase in Median Income

###### Increase in Income 25%
Between 2020 and 2021 Charlotte's median household income grew from $65,359 to $68,367, a 4.6% increase. It will be an increase of 25% from 2022 to 2027.
> Random Forest #1

```{r, scenario 2 2022}

# Simulate 10% increase in Income
data_2022_scenario2 <- data_2022 %>% 
  mutate(MedHHInc = MedHHInc * 1.25)

# Make Predictions with the New Scenario Data
predictions_2022_scenario2 <- predict(RF_model, newdata = data_2022_scenario2)

# Add Predictions to the Data
data_2022_scenario2$Predicted_Avg_ridership <- predictions_2022_scenario2

# Format Predicted Avg Ridership as Decimal Numbers
data_2022_scenario2$Predicted_Avg_ridership <- sprintf("%.2f", predictions_2022_scenario2)

# Convert Data to an sf Object
data_2022_scenario2_sf <- st_as_sf(data_2022_scenario2, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
data_2022_scenario2_sf$Predicted_Avg_ridership <- as.numeric(data_2022_scenario2_sf$Predicted_Avg_ridership)

# Plot the Map with Correct Fill Scale
ggplot(data_2022_scenario2_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Predicted_Avg_ridership), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Predicted Avg Ridership with \n10% Increase in Median Income")

```


> Random Forest #2

```{r, scenario 2 2022}

RF2_table_2022_scenario2 <- RF2_table_sf %>% 
  mutate(MedHHInc = MedHHInc * 1.25)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario2 <- predict(RF_model2, newdata = RF2_table_2022_scenario2)

# Add Predictions to the Data
RF2_table_2022_scenario2$Predicted_Avg_ridership <- predictions2_2022_scenario2

# Add Monthly Prediction Column
RF2_table_2022_scenario2 <- RF2_table_2022_scenario2 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario2$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario2)

# Convert Data to an sf Object
RF2_table_2022_scenario2_sf <- st_as_sf(RF2_table_2022_scenario2, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario2_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario2_sf$Monthly_prediction)


result <- RF2_table_2022_scenario2_sf[RF2_table_2022_scenario2_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario II", color = "Predicted Avg Ridership with \n25% Increase in Median Income")+
  theme(legend.position = "bottom")

#file_path <- "scenario2underperforming.geojson"
#st_write(result, file_path, driver = "GeoJSON")

```



###### Increase in Median Income 50%

> Random Forest #1

```{r}

# Simulate 10% increase in Income
data_2022_scenario2a <- data_2022 %>% 
  mutate(MedHHInc = MedHHInc * 1.5)

# Make Predictions with the New Scenario Data
predictions_2022_scenario2a <- predict(RF_model, newdata = data_2022_scenario2a)

# Add Predictions to the Data
data_2022_scenario2a$Predicted_Avg_ridership <- predictions_2022_scenario2a

# Format Predicted Avg Ridership as Decimal Numbers
data_2022_scenario2a$Predicted_Avg_ridership <- sprintf("%.2f", predictions_2022_scenario2a)

# Convert Data to an sf Object
data_2022_scenario2a_sf <- st_as_sf(data_2022_scenario2a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
data_2022_scenario2a_sf$Predicted_Avg_ridership <- as.numeric(data_2022_scenario2a_sf$Predicted_Avg_ridership)

# Plot the Map with Correct Fill Scale
ggplot(data_2022_scenario2a_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Predicted_Avg_ridership), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Predicted Avg Ridership with \n50% Increase in MedHHInc")

```


> Random Forest #2

```{r, scenario 2a 2022}

# Simulate 10% increase in PopDensity
RF2_table_2022_scenario2a <- RF2_table_sf %>% 
  mutate(MedHHInc = MedHHInc * 1.5)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario2a <- predict(RF_model2, newdata = RF2_table_2022_scenario2a)

# Add Predictions to the Data
RF2_table_2022_scenario2a$Predicted_Avg_ridership <- predictions2_2022_scenario2a

# Add Monthly Prediction Column
RF2_table_2022_scenario2a <- RF2_table_2022_scenario2a %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario2a$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario2a)

# Convert Data to an sf Object
RF2_table_2022_scenario2a_sf <- st_as_sf(RF2_table_2022_scenario2a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario2a_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario2a_sf$Monthly_prediction)


result <- RF2_table_2022_scenario2a_sf[RF2_table_2022_scenario2a_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario II", color = "Predicted Avg Ridership with \n50% Increase in Median Income")+
  theme(legend.position = "bottom")


#file_path <- "scenario2aunderperforming.geojson"
#st_write(result, file_path, driver = "GeoJSON")



```



##### Scenario 3: Add 2 Proposed Light Rail Stops as Feature


```{r, fig.height=8, fig.width=16}

# Rename OBJECTID_1 column to OBJECTID in red_stops
red_stops <- red_stops %>% rename(OBJECTID = OBJECTID_1)
red_stops <- red_stops %>% rename(NAME = Name)
red_stops <- select(red_stops, -Status)


# Add a unique identifier to silver_stops
silver_stops <- select(silver_stops, -c(Phase, parkRide, GeoID))
silver_stops <- silver_stops %>% rename(NAME = Name)

# Merge the datasets based on OBJECTID
proposed_line_merged_data <- rbind(existing_transit_stops, red_stops)
proposed_line_merged_data <- rbind(proposed_line_merged_data, silver_stops)

# Extract X and Y
#proposed_line_merged_data <- st_coordinates(proposed_line_merged_data)
#proposed_line_merged_data <- as.data.frame(proposed_line_merged_data)
#names(proposed_line_merged_data) <- c("x", "y")

# Assuming your dataset is named "stops_data"
proposed_line_merged_data_sf <- st_as_sf(proposed_line_merged_data, coords = c("geometry.x", "geometry.y"))


# Plot the density of points

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = proposed_line_merged_data, color = "black", size = .5, alpha = 0.5) +
  stat_density_2d(data = proposed_line_merged_data_sf, 
                  aes(x = st_coordinates(geometry)[, 1], 
                      y = st_coordinates(geometry)[, 2],
                      fill = ..level.., alpha = ..level..),
                  size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "#737373", high = "#0077B6", name = "Density") +  # Adjust color gradient
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Proposed Lines") +
  mapTheme()
  

```


```{r}

RF2_table_sf <- na.omit(RF2_table_sf)

data_2022_s1 <-
  RF2_table_sf %>% 
    mutate(
      proposed_line_nn1 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 1),
      
      proposed_line_nn2 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 2), 
      
      proposed_line_nn3 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 3), 
      
      proposed_line_nn4 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 4), 
      
      proposed_line_nn5 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 5)) 

```


```{r}
data_2022_s1 <- subset(data_2022_s1, select = -c(transit_stops_nn1, transit_stops_nn2, transit_stops_nn3, transit_stops_nn4, transit_stops_nn5))

# Rename columns using base R
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn1"] <- "transit_stops_nn1"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn2"] <- "transit_stops_nn2"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn3"] <- "transit_stops_nn3"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn4"] <- "transit_stops_nn4"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn5"] <- "transit_stops_nn5"

```



> Random Forest Model #1

```{r} 
library(randomForest)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index_s1 <- sample(1:nrow(data_2022_s1), 0.7 * nrow(data_2022_s1)) 
train_data_RF_s1 <- data_2022_s1[train_index_s1, ]
train_data_RF_s1 <- na.omit(train_data_RF_s1)
test_data_RF_s1 <- data_2022_s1[-train_index_s1, ]
test_data_RF_s1 <- na.omit(test_data_RF_s1)

# Fit the Random Forest model
RF_model_s1 <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5 + Month + PopDensity_km2 +  proposed_line_nn1 + proposed_line_nn2 + proposed_line_nn3 + proposed_line_nn4 + proposed_line_nn5, 
  data = train_data_RF_s1)


RF_testing_s1 <- test_data_RF_s1 %>%
  mutate(Predictions_RF_s1 = predict(RF_model_s1, newdata = test_data_RF_s1),
Error_RF_s1 = Predictions_RF_s1 - test_data_RF_s1$Avg_ridership,
AbsError_RF_s1 = abs(Predictions_RF_s1 - test_data_RF_s1$Avg_ridership),
MAE_RF_s1 = mean(abs(Error_RF_s1)),
APE_RF_s1 = (abs(Predictions_RF_s1 - test_data_RF_s1$Avg_ridership)) / Predictions_RF_s1)

RF_model_s1
```



```{r}
# Extract actual and predicted values
actual_s1 <- test_data_RF_s1$Avg_ridership
predicted_s1 <- predict(RF_model_s1, newdata = test_data_RF_s1)

# Create a scatter plot of actual vs. predicted values
plot(actual_s1, predicted_s1, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


```{r}
RF_testing_s1_sf <- st_as_sf(RF_testing_s1)

ggplot(RF_testing_s1_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF_s1), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


```{r}
underperforming_stops_predicted_s1 <- RF_testing_s1[RF_testing_s1$Predictions_RF_s1 <= 10, ]
underperformance_counts_predicted_s1 <- table(underperforming_stops_predicted_s1$"Stop ID")

# Top 50 Stops for 2023
top_50_stops_predicted_s1 <- names(sort(underperformance_counts_predicted_s1, decreasing = TRUE)[1:50])
underperforming_top_50_predicted_s1 <- underperforming_stops_predicted_s1[underperforming_stops_predicted_s1$"Stop ID" %in% top_50_stops_predicted_s1, ]

# Order
underperforming_top_50_predicted_s1$"Stop ID" <- factor(underperforming_top_50_predicted_s1$"Stop ID", levels = rev(top_50_stops_predicted_s1))

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, coords = c("Longitude", "Latitude"), crs = 4326)

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, wkt = "geometry", crs = 4326)

#underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)




underperforming_top_50_predicted_s1$geometry <- st_transform(underperforming_top_50_predicted_s1$geometry, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_predicted_s1 <- ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_predicted_s1, aes(geometry = geometry, colour = q5(Predictions_RF_s1)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = c("1", "2", "3", "4", "5"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Lowest Predicted Avg Ridership",
       fill = "Predictions_RF_s1") +
  mapTheme()

print(map_plot_top_50_predicted_s1)

```


> Random Forest Model #2

```{r, scenario 2a 2022}


# Make Predictions with the New Scenario Data
predictions2_2022_scenario3 <- predict(RF_model2, newdata = data_2022_s1)

# Add Predictions to the Data
data_2022_s1$Predicted_Avg_ridership <- predictions2_2022_scenario3

# Add Monthly Prediction Column
data_2022_s1 <- data_2022_s1 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario2a$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario3)

# Convert Data to an sf Object
RF2_table_2022_scenario3_sf <- st_as_sf(data_2022_s1, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario3_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario3_sf$Monthly_prediction)


result <- RF2_table_2022_scenario3_sf[RF2_table_2022_scenario3_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = silver_routes, color = "purple", size = 7, alpha = 0.5) +
  geom_sf(data = red_routes,color = "#FF6347", size = 7, alpha = 0.5) +
  geom_sf(data = blue_routes, color = "#4682B4", size = 7, alpha = 0.5) +
  geom_sf(data = gold_routes,color = "#FFD700", size = 7, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario III", color = "Predicted Avg Ridership \nwith Proposed Transit Lines Added")+
  theme(legend.position = "bottom")


#file_path <- "scenario3underperforming.geojson"
#st_write(result, file_path, driver = "GeoJSON")


```



```{r}
RF2_table_sf <- na.omit(RF2_table_sf)

RF2_table_s1 <-
  RF2_table_sf %>% 
    mutate(
      proposed_line_nn1 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 1),
      
      proposed_line_nn2 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 2), 
      
      proposed_line_nn3 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 3), 
      
      proposed_line_nn4 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 4), 
      
      proposed_line_nn5 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 5))
```


```{r} 
library(randomForest)

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index2_s1 <- sample(1:nrow(RF2_table_s1), 0.7 * nrow(RF2_table_s1)) 
train_data_RF2_s1 <- RF2_table_s1[train_index2_s1, ]
train_data_RF2_s1 <- na.omit(train_data_RF2_s1)
test_data_RF2_s1 <- RF2_table_s1[-train_index2_s1, ]
test_data_RF2_s1<- na.omit(test_data_RF2_s1)


# Fit the Random Forest model
RF_model2_s1 <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + Month + PopDensity_km2 + proposed_line_nn1 + proposed_line_nn2 + proposed_line_nn3 + proposed_line_nn4 + proposed_line_nn5+ school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center, 
  data = train_data_RF2_s1)


RF_testing2_s1 <- test_data_RF2_s1 %>%
  mutate(Predictions_RF2_s1 = predict(RF_model2_s1, newdata = test_data_RF2_s1),
Error_RF2_s1 = Predictions_RF2_s1 - test_data_RF2_s1$Avg_ridership,
AbsError_RF2_s1 = abs(Predictions_RF2_s1 - test_data_RF2_s1$Avg_ridership),
MAE_RF2_s1 = mean(abs(Error_RF2_s1)),
APE_RF2_s1 = (abs(Predictions_RF2_s1 - test_data_RF2_s1$Avg_ridership)) / Predictions_RF2_s1)

RF_model2_s1
```


```{r}
# Extract actual and predicted values
actual_RF2_s1 <- test_data_RF2_s1$Avg_ridership
predicted_RF2_s1 <- predict(RF_model2_s1, newdata = test_data_RF2_s1)

# Create a scatter plot of actual vs. predicted values
plot(actual_RF2_s1, predicted_RF2_s1, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


```{r}
RF_testing2_s1_sf <- st_as_sf(RF_testing2_s1)

ggplot(RF_testing2_s1_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF2_s1), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```

```{r}
result <- RF_testing2_s1_sf[RF_testing2_s1_sf$Month == 'January', ]
result <- result[result$Predictions_RF2_s1 <= 10, ]

ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = result, aes(color = Predictions_RF2_s1), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario III", color = "Predicted Avg Ridership with \nProposed Light Rail Routes")

```


```{r}
underperforming_stops_predicted_RF2_s1 <- RF_testing2_s1[RF_testing2_s1$Predictions_RF2_s1 <= 10, ]
underperformance_counts_predicted_RF2_s1 <- table(underperforming_stops_predicted_RF2_s1$"Stop ID")

# Top 50 Stops for 2023
top_50_stops_predicted_RF2_s1 <- names(sort(underperformance_counts_predicted_RF2_s1, decreasing = TRUE)[1:50])
underperforming_top_50_predicted_RF2_s1 <- underperforming_stops_predicted_RF2_s1[underperforming_stops_predicted_RF2_s1$"Stop ID" %in% top_50_stops_predicted_RF2_s1, ]

# Order
underperforming_top_50_predicted_RF2_s1$"Stop ID" <- factor(underperforming_top_50_predicted_RF2_s1$"Stop ID", levels = rev(top_50_stops_predicted_RF2_s1))

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, coords = c("Longitude", "Latitude"), crs = 4326)

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, wkt = "geometry", crs = 4326)

#underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)




underperforming_top_50_predicted_RF2_s1$geometry <- st_transform(underperforming_top_50_predicted_RF2_s1$geometry, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_predicted_RF2_s1 <- ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_predicted_RF2_s1, aes(geometry = geometry, colour = q5(Predictions_RF2_s1)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = c("1", "2", "3", "4", "5"),
                     name = "Quintile\nBreaks") +
  labs(title = "50 Bus Stops with Lowest Predicted Avg Ridership",
       fill = "Predictions_RF_s1") +
  mapTheme()

print(map_plot_top_50_predicted_RF2_s1)

```
