---
title: "Combined Markdown"
author: "Teresa Chang, Roshini Ganesh, Jingyi Li, Jonathan Manurung"
date: "2024-02-27"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

Link to our web app: https://roshiniganesh.shinyapps.io/MUSA-Charlotte-ShinyApp/


### 1. Introduction

In 2021, the Charlotte Area Transit System (CATS) embarked on 'The Bus Priority Study' as part of the city's larger Strategic Mobility Plan, aiming to enhance bus trips' speed, reliability, and convenience for riders. To achieve this, CATS set out to understand recent dynamics, explore accessibility, and examine network efficiencies.

#### 1.1 CATS identified three primary objectives:

1. Understanding recent bus ridership dynamics involved 
2. Exploring thepresent reality of bus accessibility, considering ridership trends and demographic distributions.
3. Examining network efficiencies and possibilities 

#### 1.2 Actions required related to objectives
Identify underperforming bus stops
Study demographic, spatial, and economic factors influencing ridership and performance  
Alter current network of stops and routes  to optimize transit services.


### 2. Use case
Our team will support CATS transportation planners in improving bus line efficiency by analyzing historical ridership trends per bus stop to identify and predict underperforming stops and routes. 

With this, we will propose alternate network options through a web-based dashboard that presents predictions with the context of demographic, spatial, financial factors of influence. 


```{r setup, include=FALSE, cache=TRUE}
# You can set some global options for knitting chunks

knitr::opts_chunk$set(echo = TRUE)

# Load some libraries
install.packages("rasterVis")

library(tidyverse)
library(sf)
library(spdep)
library(caret)
library(ckanr)
library(FNN)
library(grid)
library(gridExtra)
library(ggcorrplot) # plot correlation plot
library(corrr)      # another way to plot correlation plot
library(kableExtra)
library(jtools)     # for regression model plots
library(ggstance) # to support jtools plots
library(ggpubr)    # plotting R^2 value on ggplot point scatter
library(broom.mixed) # needed for effects plots
library(tidycensus)
library(kableExtra)
library(stargazer)
library(readxl)
library(zoo)
library(cowplot)
library(patchwork)
library(randomForest)
library(caret)
library(rasterVis)

# functions and data directory

root.dir = "https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/"
source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c")
palette7 <- c("white","#edf8fb","#ccece6","#99d8c9","#2ca25f","#006d2c", "black")
red_palette <- c("#FFD3D3", "#FFA8A8", "#FF7D7D", "#FF5252", "#FF0000")
positive_palette <- c("darkgreen", "green", "yellow", "#FFAB00", "red")
negative_palette <- c("blue", "lightblue", "grey80", "grey60", "grey40")

positive_paletteaccent <- c("#006d2c", "#2ca25f", "#FFCB62", "#FFAB00", "#C60500")
negative_paletteaccent <- c("#0077B6", "#7ACFD3", "grey80", "grey60", "grey40")

palettebase5 <- c("#edf8fb","#7ACFD3","#3AA8AE","#2D8388","#005357")
palettebase7 <- c("#edf8fb","#7ACFD3","#54C1C6","#3AA8AE","#2D8388","#016F74","#005357")
palettebase <- c("#7ACFD3","#67C8C8","#54C1C6","#41BAC0","#3AA8AE","#33969B","#2D8388","#016F74","#005357")
paletteaccent <- c("#46AFB4","#249DA2","#01939A","#017278","#005A5E","#FFCB62","#FFBD37","#FFAB00","#C68500","#9C6900","#FF6662","#FF3C37","#FF0700","#C60500","#9C0400")

red_paletteaccent <- c("#FFD3D3","#FFA8A8","#FF0700","#C60500","#9C0400")
yellow_paletteaccent <- c("#FFE6A7","#FFBD37","#FFAB00","#C68500","#9C6900")

census_api_key("ab9309f9cc70c0e1895e7166c3ca981c40cf0331", overwrite = TRUE , install = TRUE)
```

### 3. Data 

#### 3.1 Datasets considered

To analyze and forecast performance trends, we have used ridership data from CATS for the years between 2017 and 2023. Additionally we have identified potential demographic, spatial, and economic parameters that may influence our predictor variable that is ridership. 

#### 3.2 Limitations of dataset

On initial review, certain limitations of the data present themselves. These include:
1. No hourly ridership data is available precluding us from analyzing peak hour trends
2. Weekend vs weekday ridership is not available which leads to the assumption of similar usage patterns of weekends, weekdays, and holidays
3. Data on stops added or removed in these years of study is not explicit 
4. Only stop locations are made available which means route locations will have to be sourced and spatially joined. This allows us to join only the most recent route data for these stops.
5. The dataset for 2023 is incomplete and only includes data for months January to November.


```{r cars, warning=FALSE, message=FALSE, include=FALSE, cache=TRUE}
stops<-read_csv('AllDat.csv')
bus_routes <- st_read("Data/Bus_Routes.geojson")%>%
  st_transform('EPSG:3358')
school <- st_read("Data/Schools.geojson")%>%
  st_transform('EPSG:3358')

parks <- st_read("Data/Parks.geojson")%>%
  st_transform('EPSG:3358')

groceries <-
  st_read("Data/Grocery_Stores.geojson")%>%
  st_transform('EPSG:3358')

shoppingcen <-
  st_read("Data/Existing_Shopping_Centers.geojson")%>%
  st_transform('EPSG:3358')

policeoffice <-
  st_read("Data/CMPD_Police_Division_Offices.geojson") %>%
  st_transform('EPSG:3358')

# Create a vector of years
years <- c(2017, 2018, 2019, 2020, 2021, 2022)

# Loop through the years to load variables
variables_list <- lapply(years, function(year) {
  load_variables(year, "acs5", cache = TRUE)
})

# Combine the variables from different years into a single data frame
all_variables <- do.call(rbind, variables_list)

hlt <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.geojson")

# 2017 census
dat2017 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2017, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2018
dat2018 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2018, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2018") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

#census 2019
dat2019 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2019, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2019") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2020
dat2020 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2020, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale)

# census 2021
dat2021 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2021, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2021") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 

# census 2022
dat2022 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E","B08201_001E",
                        "B08201_002E","B08201_003E",
                        "B08201_004E","B08201_005E",
                        "B08201_006E","B08201_007E",
                        "B08301_002E","B08301_016E",
                        "B08301_017E","B08301_018E",
                        "B08301_019E"), 
          year=2022, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('EPSG:3358') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Workfromhome = B08111_031E,
         Total = B08201_001E,
         car0 = B08201_002E,
         car1 = B08201_003E,
         car2 = B08201_004E,
         car3 = B08201_005E,
         car4 = B08201_006E,
         car5more = B08201_007E,
         Commutetowork = B08301_001E,
         Workwithbus = B08301_011E,
         Workbycar = B08301_002E,
         Workbytaxi = B08301_016E,
         Workbymotor = B08301_017E,
         Workbybike = B08301_018E,
         Workbywalk = B08301_019E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(PopDensity_km2 = ifelse(TotalPop > 0, TotalPop / (st_area(.) / 1e6), 0),
         TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Commutetowork > 0, Workwithbus / Commutetowork, 0),
         pctworkfromhome = ifelse(Commutetowork > 0, Workfromhome / Commutetowork, 0),
         pctworkbycar = ifelse(Commutetowork > 0, Workbycar / Commutetowork, 0),
         pctworkbytaxi = ifelse(Commutetowork > 0, Workbytaxi / Commutetowork, 0),
         pctworkbymotor = ifelse(Commutetowork > 0, Workbymotor / Commutetowork, 0),
         pctworkbybike = ifelse(Commutetowork > 0, Workbybike / Commutetowork, 0),
         pctworkbywalk = ifelse(Commutetowork > 0, Workbywalk / Commutetowork, 0),
         pctnocar = ifelse(Total > 0, car0 / Total, 0),
         pct1car = ifelse(Total > 0, car1 / Total, 0),
         pct2car = ifelse(Total > 0, car2 / Total, 0),
         pct3car = ifelse(Total > 0, car3 / Total, 0),
         pct4car = ifelse(Total > 0, car4 / Total, 0),
         pct5car = ifelse(Total > 0, car5more / Total, 0),
         year = "2022") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 


```

#### 3.3 Data Wrangling

```{r pressure, , warning=FALSE, message=FALSE, include=FALSE, echo=FALSE, cache=TRUE}
#feature engineering
school.sf <-
  school %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

parks.sf <-
  parks %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

groceries.sf <-
  groceries %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

shoppingcen.sf <-
  shoppingcen %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

policeoffice.sf <-
  policeoffice %>%
    dplyr::select(geometry) %>%
    na.omit() %>%
    filter(!st_is_empty(.))%>%
    st_as_sf(coords = c("Long", "Lat"), crs = "EPSG:3358")%>%
    distinct()

silver_stops <- st_read("Data/LYNX_Silver_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

silver_routes <- st_read("Data/LYNX_Silver_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')

blue_stops <-
  st_read("Data/LYNX_Blue_Line_Stations.geojson")%>%
  st_transform('EPSG:3358')

blue_routes <-
  st_read("Data/LYNX_Blue_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

gold_stops <-
  st_read("Data/LYNX_Gold_Line_Stops.geojson")%>%
  st_transform('EPSG:3358')

gold_routes <-
  st_read("Data/LYNX_Gold_Line_Route.geojson")%>%
  st_transform('EPSG:3358')

red_stops <-
  st_read("Data/LYNX_Red_Line_Stations_Proposed.geojson")%>%
  st_transform('EPSG:3358')

red_routes <-
  st_read("Data/LYNX_Red_Line_Route_Proposed.geojson")%>%
  st_transform('EPSG:3358')


allTracts <- rbind(
  dat2017,
  dat2018,
  dat2019,
  dat2020,
  dat2021,
  dat2022)
sf_allTracts <- st_as_sf(allTracts, wkt = "geometry")

bus_data <- stops %>% mutate(`Stop ID` = sprintf("%05s", `Stop ID`))
busstop_sf <- st_as_sf(bus_data, coords = c("Longitude", "Latitude"), crs = 4326)
catsstop_sf <- st_as_sf(stops, coords = c("Longitude", "Latitude"), crs = 2264)  # Assuming NAD 83/North Carolina (ftUS)
#Reproject catsstop_sf to WGS 84 (EPSG:4326)
catsstop_sf <- st_transform(catsstop_sf, 4326)
busstop_sf <- st_transform(busstop_sf, 3358)
catsstop_sf <- st_transform(catsstop_sf, 3358)
routes_sf <- st_transform(bus_routes, 3358)
allTracts_sf <- st_transform(allTracts, 3358)

bus_stops.sf <-
  stops %>%
    st_as_sf(coords = c("Longitude", "Latitude"), crs = "EPSG:3358")%>%
    distinct()



```

Initial Visualization of Board and Alight from APC Data

#### 3.4 APC data

```{r, warning = FALSE, message=FALSE, cache=TRUE}
ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Board), size = .3) +
  scale_color_gradient(low = "#FFAB00", high = "darkgreen") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Board Value")

ggplot(busstop_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = Alight), size = .3) +
  scale_color_gradient(low = "#FFAB00", high = "darkgreen") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Alight Value")

```

### 4. Initial Trends

1. Ridership does not seem to vary too much over the years across all stops based on the given CATS dataset. However, demographic data from ACS 2022 suggests that there has been a steep decline in the percentage of people who commute to work by bus over the last 7 years. This may be attributed to more wealthy people working from home along with more economically disadvantaged people choosing to take the bus. 
2. The global trend suggests that bus ridership during the pandemic did not drop significantly. We suspect that the consistent ridership profile may be due to errors in recording the data for pandemic years. 

#### 4.1 Performance before and after the COVID-19 pandemic

To investigate the nature of performance before and after pandemic years further, the difference between average ridership is mapped pre Covid (2017- 2020/3) and post Covid (2020/3 - 2023). This mapping exercise indicates that ridership has decreased in most bus stops post the pandemic. Additionally, some stops show a hundred percent decline. These are filtered out because it may indicate that the stop is no longer in service.

Next, stops with an increase in ridership are separated from stops with a decrease and the results are mapped. This suggests that a majority of stops in the city center have experienced a decrease in ridership. This spatial clustering of underperforming stops suggests areas for targeted intervention.


```{r, warning=FALSE, message=FALSE, include=FALSE, cache=TRUE}
# threshold: date of covid bgan
threshold_date <- as.Date("2020-03-01") 

stops_data <- stops %>%
  mutate(Numeric_Month = match(Month, month.name)) %>%
  mutate(Readable_Date = as.yearmon(paste(Year, sprintf("%02d", Numeric_Month)), "%Y %m")) %>%
  mutate(Covid_Status = ifelse(as.Date(Readable_Date) >= threshold_date, "Post", "Pre"))%>%
  rename(Stop_Name = `Stop Name`)%>%
  rename(Stop_ID = `Stop ID`)

# Calculate the increase/decrease
average_board <- stops_data %>%
  group_by(Stop_Name, Numeric_Month, Covid_Status) %>%
  summarize(Avg_Board = mean(Avg_Board, na.rm = TRUE))

average_board_pivoted <- pivot_wider(average_board, names_from = Covid_Status, values_from = Avg_Board)

average_board_pivoted <- average_board_pivoted %>%
  mutate(Difference = (Post - Pre)/Pre *100 ) %>%
  select(Stop_Name, Numeric_Month, Pre, Post, Difference)

average_board_pivoted <- na.omit(average_board_pivoted)
# join the longitude and latitude data back and select the first one
average_board_pivoted <- average_board_pivoted %>%
  left_join(stops_data %>% select(Stop_Name, Longitude, Latitude) %>% distinct(Stop_Name, .keep_all = TRUE), by = "Stop_Name")

# Adjust the month as needed
selected_month <- 1  
selected_data <- average_board_pivoted %>%
  filter(Numeric_Month == selected_month)

stops_sf <- st_as_sf(selected_data, coords = c("Longitude", "Latitude"), crs = 4326)
stops_sf <- st_transform(stops_sf, crs = 3358)  # NAD83 North Carolina

```

##### 4.1.1 All stops from 2017\~2023

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
palettebase5_rev <- rev(palettebase5)

#ggplot() +
#  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
#  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
#  geom_sf(data = stops_sf , aes(colour = q5(Difference)), 
#          show.legend = "point", size = 1, alpha= 0.5) +
#  scale_colour_manual(values = palettebase5_rev,
#                   labels=qBr(stops_sf ,"Difference"),
#                   name="Quintile\nBreaks (%)") +
#  labs(title = paste("Difference (%) - Month:", selected_month),
#       fill = "Difference (Post - Pre)") +
#  mapTheme()

filtered_stops_sf <- stops_sf[stops_sf$Difference != -100, ]

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = filtered_stops_sf , aes(colour = q5(Difference)), 
          show.legend = "point", size = 1, alpha= 0.5) +
  scale_colour_manual(values = palettebase5_rev,
                   labels=qBr(filtered_stops_sf ,"Difference"),
                   name="Quintile\nBreaks (%)") +
  labs(title = paste("Difference (%) (-100% excluded)   - Month:", selected_month),
       fill = "Difference (Post - Pre)") +
  mapTheme()
```

##### 4.1.2 Increase vs Decrease

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
stops_sf$PositiveDifference <- stops_sf$Difference > 0
positive_stops_sf <- stops_sf[stops_sf$PositiveDifference,]
negative_stops_sf <- stops_sf[!stops_sf$PositiveDifference,]

yellow_paletteaccent_rev <- rev(yellow_paletteaccent)

# positive and negative differences
positive_plot <- ggplot(positive_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = red_paletteaccent,
                     labels = qBr(positive_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Positive Increase (%) - Month:", selected_month),
       fill = "Positive Difference (Post - Pre)") +
  mapTheme()

negative_plot <- ggplot(negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()

#combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)
#print(combined_plot)

# Exclude -100%
filtered_negative_stops_sf <- negative_stops_sf[negative_stops_sf$Difference != -100, ]

negative_plot <- ggplot(filtered_negative_stops_sf, aes(colour = q5(Difference))) +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(size = 1) +
  scale_color_manual(values = yellow_paletteaccent_rev,
                     labels = qBr(filtered_negative_stops_sf, "Difference"),
                     name = "Quintile\nBreaks (%)") +
  labs(title = paste("Negative Decrease (%) (-100% excluded) - Month:", selected_month),
       fill = "Negative Difference (Post - Pre)") +
  mapTheme()
combined_plot <- plot_grid(positive_plot, negative_plot, ncol = 2)

print(combined_plot)
```

> From the plotted maps, there is a majority of stops in Charlotte’s city center that have experienced a decrease in ridership before and after the COVID-19 pandemic



#### 4.2 Top 50 underperforming stops from 2017 to 2023

Between 2017-2023, there is a uniform trend of underperformance in certain underperforming stops. Other underperforming stops exhibit a growing trend of underperformance post the pandemic. About half of these stops exhibit spatial clustering patterns and are located along a single route.


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
# filter underperforming stops
underperforming_stops <- stops_data[(stops_data$Board + stops_data$Alight) <= 10, ]
underperformance_counts <- table(underperforming_stops$Stop_ID)


top_50_stops <- names(sort(underperformance_counts, decreasing = TRUE)[1:50])
underperforming_top_50 <- underperforming_stops[underperforming_stops$Stop_ID %in% top_50_stops, ]
underperforming_top_50 <- underperforming_top_50 %>%
  mutate(Total_Board_Alight = Board + Alight)

# Order
underperforming_top_50$Stop_ID <- factor(underperforming_top_50$Stop_ID, levels = rev(top_50_stops))

stacked_bar_plot <- ggplot(underperforming_top_50, aes(x = Stop_ID, fill = factor(Year))) +
  geom_bar(stat = "count", position = "stack", color = "darkgrey" ) +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = palettebase7) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  # Rotate x-axis labels
## print(stacked_bar_plot) 

underperforming_top_50_all_sf <- st_as_sf(underperforming_top_50, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_all_sf <- st_transform(underperforming_top_50_all_sf, crs = 3358)  # NAD83 North Carolina


map_plot_top_50 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_all_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_all_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts (2017 - 2023)",
       fill = "Avg_Board + Avg_Alight") +
  mapTheme()
## print(map_plot_top_50) 

combined_plot <- stacked_bar_plot + map_plot_top_50
print(combined_plot)
```

#### 4.3 Top 50 underperforming stops in 2022

The underperformance and spatial clustering trends in 2023 are consistent with the overall trend.

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
# Underperforming Stops in 2022
underperforming_stops_2022 <- stops_data[(stops_data$Board + stops_data$Alight) <= 10 & stops_data$Year == 2022, ]
underperformance_counts_2022 <- table(underperforming_stops_2022$Stop_ID)

# Top 30 Stops for 2023
top_50_stops_2022 <- names(sort(underperformance_counts_2022, decreasing = TRUE)[1:50])
underperforming_top_50_2022 <- underperforming_stops_2022[underperforming_stops_2022$Stop_ID %in% top_50_stops_2022, ]


# Order
underperforming_top_50_2022$Stop_ID <- factor(underperforming_top_50_2022$Stop_ID, levels = rev(top_50_stops_2022))

bar_plot_top_50_2022 <- ggplot(underperforming_top_50_2022, aes(x = Stop_ID, fill = "2022")) +
  geom_bar(stat = "count", color = "black") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts",
       x = "Bus Stop Name",
       y = "Number of Times Underperformed",
       fill = "Year") +
  scale_fill_manual(values = c("2022" = "#7ACFD3"), name = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
##print(bar_plot_top_50_2022)

underperforming_top_50_2022 <- underperforming_top_50_2022 %>%
  mutate(Total_Board_Alight = Board + Alight)


underperforming_top_50_sf <- st_as_sf(underperforming_top_50_2022, coords = c("Longitude", "Latitude"), crs = 4326)
underperforming_top_50_sf <- st_transform(underperforming_top_50_sf, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_2022 <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_sf, aes(colour = q5(Total_Board_Alight)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = qBr(underperforming_top_50_sf, "Total_Board_Alight"),
                     name = "Quintile\nBreaks") +
  labs(title = "Top 50 Bus Stops with Highest Underperformance Counts in 2022",
       fill = "Total_Board_Alight") +
  mapTheme()
##print(map_plot_top_50_2022)

combined_plot <- bar_plot_top_50_2022 + map_plot_top_50_2022
print(combined_plot)
```


> Some stops exhibit a growing trend of underperformance post the pandemic.About half of these stops exhibit spatial clustering patterns and are located along a single route. The underperformance and clustering trends in 2022 are consistent with the overall trend in the city center area.


```{r, warning=FALSE, message=FALSE, fig.height=10, fig.width=8, cache=TRUE}
##print(map_plot_top_50) 
##print(map_plot_top_50_2022)

##print(stacked_bar_plot) 
##print(bar_plot_top_50_2022)
```

### 5. Spatial factors:

#### 5.1 Existing and proposed transit lines and stations

Other existing and proposed transit stops and routes are considered to study the relationship between infrastructure and underperforming stops, if any. If significant, the distance of bus stops to other transit stops would also serve as variables in the model.


```{r, warning=FALSE, message=FALSE, cache=TRUE}
transit <- ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, aes(color = "Bus Route"), size = 5, alpha = 0.5) +
  geom_sf(data = blue_routes, aes(color = "Blue Line"), size = 8) +
  geom_sf(data = blue_stops, aes(color = "Blue Line"), size = 2) +
  geom_sf(data = gold_routes, aes(color = "Gold Line"), size = 8) +
  geom_sf(data = gold_stops, aes(color = "Gold Line"), size = 2) +
  geom_sf(data = silver_routes, aes(color = "Silver Line"), size = 8, linetype = "dashed") +
  geom_sf(data = silver_stops, aes(color = "Silver Line"), size = 2, shape = 21) +
  geom_sf(data = red_routes, aes(color = "Red Line"), size = 8, linetype = "dashed") +
  geom_sf(data = red_stops, aes(color = "Red Line"), size = 2, shape = 21) +
  
  labs(title = "Existing and Proposed Transit Lines and Stops") +
    scale_color_manual(name = "Lines",
                     values = c("Bus Route" = "darkgrey",
                                "Blue Line" = "#4682B4", 
                                "Gold Line" = "#FFD700", 
                                "Silver Line" = "purple", 
                                "Red Line" = "#FF6347"),
                     labels = c("Blue Line","Bus Routes", "Gold Line", "Proposed Red Line", "Proposed Silver Line")) +

  mapTheme()

print(transit)

```

#### 5.2 Density of Amenities

Most amenities are clustered within the city center and expanded to the south, presenting an inequitable distribution. For model input, the distance to the amenities would serve as variables as well

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
A <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = school, color = "darkgrey", size = .5, alpha = 0.3) +
  stat_density2d(data = data.frame(st_coordinates(school.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFE6A7", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Schools") +
  mapTheme()

B <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = parks, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(parks.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#7ACFD3", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Parks") +
  mapTheme()

C <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = groceries, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(groceries.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FFAB00", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Grocery Stores") +
  mapTheme()

D <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = policeoffice, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(policeoffice.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#0077B6", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Police Offices") +
  mapTheme()

E <- ggplot() + geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = shoppingcen, color = "darkgrey", size = .5, alpha = 0.5) +
  stat_density2d(data = data.frame(st_coordinates(shoppingcen.sf)), 
                 aes(X, Y, fill = ..level.., alpha = ..level..),
                 size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "grey", high = "#FF6662", name = "Density") +
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Shopping Centers") +
  mapTheme()


combined_maps <- A + B + C + D + E
#combined_maps <- combined_maps / plot_layout(nrow = 2, ncol = 3)
print(combined_maps)
```

### 6. Demographic Factors

#### 6.1 Mode Trends 

Almost all census tracts with a higher percentage of population that commute by bus to work are located close to bus stops. However, some suburban census tracts with a high percentage of population that commute by bus are far from bus stops. This may be considered as a variable that influences prediction.

##### 6.1.1 By Year and Population

```{r Trends by Year, warning=FALSE, message=FALSE, fig.width=8, fig.height=4, cache=TRUE}
a <- ggplot(allTracts, aes(x = as.factor(year), y = TotalPop, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Total Population: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

b <- ggplot(allTracts, aes(x = as.factor(year), y = TotalWorker, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Worker: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

c <- ggplot(allTracts, aes(x = as.factor(year), y = Workfromhome, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Work from Home: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)
d <- ggplot(allTracts, aes(x = as.factor(year), y = Workwithbus, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of People Using Bus as Transportation to Work: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palettebase7)

combined_plot <- grid.arrange(a,b,c,d, ncol = 4, top = c(2, 2))

print(combined_plot)

```

##### 6.1.2 Mode Share

The general trend of percentage of population commute by bus has decreased. The 2022 percentage is halved comparing to 2017 percentage. In comparison, the general trend of percentage of population commute by car has decreased as by more than 10%. Additionally, the general trend of percentage of population commute by bicycle has a subtle decrease. Over the 6 year, it decreased 0.07%. Finally, the general trend of percentage of population commute by walk has slightly increased. Its peak was during Covid-19, but post Covid-19 period, the rate drops.

```{r, warning=FALSE, message=FALSE, cache=TRUE}
allTracts$year <- as.numeric(allTracts$year)

average_pct_bus <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bus = mean(pctworkwithbus))

ggplot(average_pct_bus, aes(x = year, y = avg_pct_bus)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Bus",
       title = "Trend of Commuting by Bus from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_car <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_car = mean(pctworkbycar))

ggplot(average_pct_car, aes(x = year, y = avg_pct_car)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Car",
       title = "Trend of Commuting by Car from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

average_pct_taxi <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_taxi = mean(pctworkbytaxi))

ggplot(average_pct_taxi, aes(x = year, y = avg_pct_taxi)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Taxi",
       title = "Trend of Commuting by Taxi from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_motor <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_motor = mean(pctworkbymotor))

ggplot(average_pct_motor, aes(x = year, y = avg_pct_motor)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting with Motorcycle",
       title = "Trend of Commuting by Motorcycle from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_bike <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_bike = mean(pctworkbybike))

ggplot(average_pct_bike, aes(x = year, y = avg_pct_bike)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Bike",
       title = "Trend of Commuting by Bike from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  theme_minimal()

average_pct_walk <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_walk = mean(pctworkbywalk))

ggplot(average_pct_walk, aes(x = year, y = avg_pct_walk)) +
  geom_line(color = "black",linewidth=1.5) +
  geom_point(color = "red",size=3) +
  labs(x = "Year", 
       y = "Average Percentage of Population Commuting by Walk",
       title = "Trend of Commuting by Walk from 2017 to 2022") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  theme_minimal()

```


##### 6.1.3 Demographic Maps

The majority of Mecklenburg County population resides in suburban census tract, a trend that is consistent with the number of employed individuals in the county who also live in the suburbs. However, when examining census data on the use of public transportation to commute to work, a significant proportion is concentrated in the city center and expanded to the south. This can be correlated with census data on the prevalence of working from home, which tends to be dispersed in suburban areas situated relatively farther away from the city center


```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12, cache=TRUE}
tp_vars2 <- c("TotalPop", "TotalWorker", "Workwithbus", "Workfromhome" )
tp_titles2 <- c("Total Population ", "Total Population of Worker", "Total Population Commute with Bus", "Total Population Work from Home ")
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2017,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2017 (Census Tracts)"))

# plot 2018
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2018,aes(fill = (!!sym(i))), crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2018 (Census Tracts)"))

# plot 2019
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2019,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2019 (Census Tracts)"))

#plot 2020
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2020,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2020 (Census Tracts)"))

#plot 2021
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2021,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2021 (Census Tracts)"))

#plot 2022
mapList2 <- list()
j2 <- 1

for(i in tp_vars2){
  map2 <- ggplot() +
         geom_sf(data = dat2022,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles2[j2], caption = paste0("Figure ", as.character(j2))) +
  mapTheme()
  mapList2[[i]] <- map2
  j2<-j2+1
}

do.call(grid.arrange,c(mapList2, ncol=2, top="Demography of Mecklenburg County in 2022 (Census Tracts)"))
```

> From the plotted maps, suburban census tracts in Charlotte have more residents as well as more number of employed workers. For the people commuting to work by bus are largely concentrated in the city center of Charlotte. Meanwhile, work from home trends indicate dominant concentrations in the suburban areas.


##### 6.1.4 Spatial Reference of Car Ownership and Commute by Bus

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12, cache=TRUE}
tp_vars <- c("pctworkwithbus", "pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
tp_titles <- c("Percent of Population Commute with Bus", "Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2017,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2017 (Census Tracts)"))

# plot 2018
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2018,aes(fill = (!!sym(i))), crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2018 (Census Tracts)"))

# plot 2019
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2019,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2019 (Census Tracts)"))

#plot 2020
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2020,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2020 (Census Tracts)"))

#plot 2021
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2021,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2021 (Census Tracts)"))

#plot 2022
mapList <- list()
j <- 1

for(i in tp_vars){
  map <- ggplot() +
         geom_sf(data = dat2022,aes(fill = (!!sym(i))),crs = st_crs("EPSG:4326")) +
         scale_fill_gradient(low = "#edf8fb", high = "#005357") + 
         geom_sf(data=bus_routes, color = "#FFBD37", size = 0.5,alpha = 0.5)+
         labs(title = tp_titles[j], caption = paste0("Figure ", as.character(j))) +
    mapTheme()
  mapList[[i]] <- map
  j<-j+1
}

do.call(grid.arrange,c(mapList, ncol=2, top="Commute by Bus and Car Ownership of Mecklenburg County in 2022 (Census Tracts)"))
```


##### 6.1.5 Histogram of Car Ownership

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.height=18, fig.width=12, cache=TRUE}
car_vars <- c("pctnocar", "pct1car", "pct2car", "pct3car", "pct4car", "pct5car")
car_titles <- c("Percent of Household with No Car", "Percent of Household with 1 Car", "Percent of Household with 2 Cars", "Percent of Household with 3 Cars", "Percent of Household with 4 Cars", "Percent of Household with 5 or more Cars")
# hist 2017
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2017"))

# hist 2018
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2018"))

# hist 2019
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2019"))

# hist 2020
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2020"))

# hist 2021
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2021"))

# hist 2022
histList <- list()
j <- 1

for (i in car_vars) {
  hist <- ggplot(data = dat2017, aes(x = !!sym(i))) +
    geom_histogram(fill = "#edf8fb", color = "black") +  # Set initial color to black (can be any color)
    labs(title = car_titles[j], caption = paste0("Figure ", as.character(j))) +
    theme_minimal() +
    scale_fill_manual(values = palettebase7)  # Using scale_fill_manual to specify colors
  histList[[i]] <- hist
  j <- j + 1
}

do.call(grid.arrange,c(histList, ncol=2, top="Histogram of Car Ownership of Mecklenburg County in 2022"))
```

> From the plotted maps, suburban census tracts in Charlotte have more households with cars than household in city center. Most of those households have two or more cars


##### 6.1.6 Car Ownership Trend

```{r, warning=FALSE, message=FALSE, cache=TRUE}
average_pct_cars <- allTracts %>%
  group_by(year) %>%
  summarise(avg_pct_0cars = mean(pctnocar),
            avg_pct_1cars = mean(pct1car),
            avg_pct_2cars = mean(pct2car),
            avg_pct_3cars = mean(pct3car),
            avg_pct_4cars = mean(pct4car),
            avg_pct_5ormore_cars = mean(pct5car))

average_pct_cars_long <- pivot_longer(average_pct_cars, 
                                      cols = starts_with("avg_pct"), 
                                      names_to = "Cars", 
                                      values_to = "Percentage")

plot_list <- list()
for (i in unique(average_pct_cars_long$Cars)) {
  num_cars <- gsub("avg_pct_", "", i)
  num_cars <- gsub("cars", "car", num_cars)
  subset_data <- subset(average_pct_cars_long, Cars == i)
  plot <- ggplot(subset_data, aes(x = year, y = Percentage, group = 1)) +
    geom_line(color = "black",linewidth=1.5) +
    geom_point(color = "red",size=3) +
    labs(x = "Year", 
         y = "Average Percentage of Households",
         title = paste("Trend of Car Ownership:", num_cars)) +
    scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
    theme_minimal()
  plot_list[[i]] <- plot
}

grid.arrange(grobs = plot_list, ncol = 2, top = "Trend of Car Ownership (2017-2022)")
```

### 7. Next Steps

1.To identify most significant demographic and spatial variables for predicting underperformance
2. To present findings to clients and solicit feedback
3. To arrive at best modeling process to predict stops that are likely to underperform
4. To propose removal of and addition of stops to certain routes and examine their equity and financial impact
5. To create a web-based dashboard that allows CATS transportation planners to overlay socioeconomic, equity, and financial factors with underperformance to be able to allocate resources to stops effectively

In doing so, our aim is to build a comprehensive and robust model to optimize bus line efficiency, aligning with CATS' goal of providing faster, more reliable, and convenient transit options for Charlotte residents.


```{r, warning=FALSE, message=FALSE, cache=TRUE}
# Perform spatial join based on containment (bus stops within census tracts)
joined_data <- st_join(busstop_sf, allTracts_sf, join = st_within)
```


```{r, warning=FALSE, message=FALSE, cache=TRUE}
a# Spatial join based on containment (bus stops within census tracts)
if (st_crs(busstop_sf) != st_crs(allTracts_sf)) {
  busstop_sf <- st_transform(busstop_sf, crs = st_crs(allTracts_sf))
}

# Spatial join
joined_data <- st_join(busstop_sf, allTracts_sf, join = st_within)

# Separate dataset
data_2017 <- filter(joined_data, Year == 2017 & year == 2017)
data_2018 <- filter(joined_data, Year == 2018 & year == 2018)
data_2019 <- filter(joined_data, Year == 2019 & year == 2019)
data_2020 <- filter(joined_data, Year == 2020 & year == 2020)
data_2021 <- filter(joined_data, Year == 2021 & year == 2021)
data_2022 <- filter(joined_data, Year == 2022 & year == 2022)

# Save csv
#write.csv(data_2017, "data_2017.csv", row.names = FALSE)
#write.csv(data_2018, "data_2018.csv", row.names = FALSE)
#write.csv(data_2019, "data_2019.csv", row.names = FALSE)
#write.csv(data_2020, "data_2020.csv", row.names = FALSE)
#write.csv(data_2021, "data_2021.csv", row.names = FALSE)
#write.csv(data_2022, "data_2022.csv", row.names = FALSE)
```


```{r clean_school, cache=TRUE}
#joined_data$school.Buffer <- joined_data %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(school.sf, counter = 1),., sum) %>%
#    pull(counter)

#joined_data$policeoffice.Buffer <- joined_data %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(policeoffice.sf, counter = 1),., sum) %>%
#    pull(counter)

#data_2022$busstop.Buffer <- data_2022 %>% 
#    st_buffer(660) %>% 
#    aggregate(mutate(bus_stops.sf, counter = 1),., sum) %>%
#    pull(counter)


data_2022 <-
  data_2022 %>% 
    mutate(
      school_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 1),
      
      school_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 2), 
      
      school_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 3), 
      
      school_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 4), 
      
      school_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(school.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      policeoffice_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 1),
      
      policeoffice_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 2), 
      
      policeoffice_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 3), 
      
      policeoffice_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 4), 
      
      policeoffice_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(policeoffice.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      groceries_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 1),
      
      groceries_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 2), 
      
      groceries_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 3), 
      
      groceries_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 4), 
      
      groceries_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(groceries.sf), k = 5)) 


data_2022 <-
  data_2022 %>% 
    mutate(
      parks_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 1),
      
      parks_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 2), 
      
      parks_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 3), 
      
      parks_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 4), 
      
      parks_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(parks.sf), k = 5)) 

data_2022 <-
  data_2022 %>% 
    mutate(
      shoppingcen_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 1),
      
      shoppingcen_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 2), 
      
      shoppingcen_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 3), 
      
      shoppingcen_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 4), 
      
      shoppingcen_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(shoppingcen.sf), k = 5)) 

```


```{r, cache=TRUE}
# Rename OBJECTID_1 column to OBJECTID in blue_stops
blue_stops <- blue_stops %>% rename(OBJECTID = OBJECTID_1)
blue_stops <- select(blue_stops, c(OBJECTID, NAME))

gold_stops <- select(gold_stops, c(OBJECTID, Stop_Name))
gold_stops <- gold_stops %>% rename(NAME = Stop_Name)
# Merge the datasets based on OBJECTID
existing_transit_stops <- rbind(blue_stops, gold_stops)
existing_transit_stops <- st_as_sf(existing_transit_stops, coords = c("geometry.x", "geometry.y"))


# Plot the density of points
ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = existing_transit_stops, color = "black", size = .5, alpha = 0.5) +
  stat_density_2d(data = existing_transit_stops, 
                  aes(x = st_coordinates(geometry)[, 1], 
                      y = st_coordinates(geometry)[, 2],
                      fill = ..level.., alpha = ..level..),
                  size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "#737373", high = "#0077B6", name = "Density") +  # Adjust color gradient
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Existing Light Rail Stops") +
  mapTheme()

```

```{r, cache=TRUE}
data_2022 <-
  data_2022 %>% 
    mutate(
      transit_stops_nn1 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 1),
      
      transit_stops_nn2 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 2), 
      
      transit_stops_nn3 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 3), 
      
      transit_stops_nn4 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 4), 
      
      transit_stops_nn5 = nn_function(st_coordinates(data_2022), 
                              st_coordinates(existing_transit_stops), k = 5)) 
```

```{r, cache=TRUE}
city_center<- st_sfc(st_point(c(1692115, 5236081)), crs = 3358)
data_2022$dist_to_center <- st_distance(data_2022$geometry, city_center)
```


### 8.Modeling

#### 8.1 OLS

##### 8.1.1 Modeling OLS

We first have to have the independent variable and predictors ready, the we split the 2022 transit data into 70/30 of training and testing datasets. Then we fit a linear regression model using the training data. The model predicts Avg_ridership based on a variety of predictors including demographic data, transportation details, and nearest neighbor indices for various services.
Using the test data, we calculate predictions and evaluate the error metrics including Mean Absolute Error (MAE) and Absolute Percentage Error (APE). These metrics help in understanding the accuracy of the model predictions. The R^2 is around 0.17, which is not a optimal number for a model of high predicting power.Although some predictors demonstrate significant correlations with the independent variable, others are not showing this significance. 
```{r, cache=TRUE}
joined_data_test <- joined_data_test %>% 
             mutate(Avg_ridership = Avg_Board + Avg_Alight)
```

```{r, cache=TRUE}
data_2022 <- data_2022 %>% 
             mutate(Avg_ridership = Avg_Board + Avg_Alight)
# split traning and testing
set.seed(123) # Set seed for reproducibility
train_index <- createDataPartition(data_2022$Avg_ridership, p = 0.7, list = FALSE)
OLS_train_data <- data_2022[train_index, ]
OLS_test_data <- data_2022[-train_index, ]

#Fit regression
OLS <- lm(Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center
          , data = OLS_train_data)

# Evaluate model
OLS_testing <- OLS_test_data %>%
  mutate(
    Predictions = predict(OLS, OLS_test_data),
    Error = Predictions - OLS_test_data$Avg_ridership,
    AbsError = abs(Predictions - OLS_test_data$Avg_ridership),
    MAE = mean(abs(Error), na.rm = TRUE),
    APE = abs(Predictions - OLS_test_data$Avg_ridership) / Predictions)
summary(OLS)
```

Residual plots doesn't scatter randomly, indicating potential heteroscedasticity.

```{r, cache=TRUE}
plot(OLS)
```


##### 8.1.2 Visualize Prediction vs. Actual

We plot the actual vs. predicted ridership to visually assess the model's performance. A line of perfect prediction is added for reference. Based on our results, there are many points of predictions off the line actuality by much, therefore we do not consider the OLS model has high predictive power and accuracy.

```{r, cache=TRUE}
# Extract actual and predicted values
OLSactual <- OLS_test_data$Avg_ridership
OLSpredicted <- predict(OLS, newdata = OLS_test_data)

# Create a scatter plot of actual vs. predicted values
plot(OLSactual, OLSpredicted, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


Using the testing dataset now converted to a spatial format, we create a map to visualize the absolute error in predictions across different geographic locations.The map indicates that for some stops the absolute errors of the OLS model is larger than 300 ridership, further indicates this model does not perform well in predicting average ridership based on the predictors provided.

```{r, cache=TRUE}
OLS_testing_sf <- st_as_sf(OLS_testing)

ggplot(OLS_testing_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


#### 8.2 Random Forest Model 1 : Bus Stops
##### 8.2.1 Running the Model

Same as the OLS model, we ran a random forest model of bus stops following the same process. The output from your Random Forest regression model indicates a highly effective model for predicting Avg_ridership, as shown by the 94.95% of variance explained in the dependent variable. This high percentage suggests that the model does an excellent job of capturing the relationships and patterns within the data, which is further supported by the mean of squared residuals at 18.31362, indicating the predictions are generally close to the actual values. The model, built with 500 trees and evaluating 16 variables at each split, demonstrates robust predictive performance.

```{r, cache=TRUE} 
# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index <- sample(1:nrow(data_2022), 0.7 * nrow(data_2022)) 
train_data_RF <- data_2022[train_index, ]
train_data_RF <- na.omit(train_data_RF)
test_data_RF <- data_2022[-train_index, ]
test_data_RF<-na.omit(test_data_RF)

# Fit the Random Forest model
RF_model <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5 + Month + PopDensity_km2, 
  data = train_data_RF)


RF_testing <- test_data_RF %>%
  mutate(Predictions_RF = predict(RF_model, newdata = test_data_RF),
Error_RF = Predictions_RF - test_data_RF$Avg_ridership,
AbsError_RF = abs(Predictions_RF - test_data_RF$Avg_ridership),
MAE_RF= mean(abs(Error_RF)),
APE_RF = (abs(Predictions_RF - test_data_RF$Avg_ridership)) / Predictions_RF)

RF_model
```

We also created this Actual vs. Prediction plot, we can see the predictions are now more close to the line of perfect prediction than OLS model, which further indicates the fact that the performance of random forest model is better.

```{r, cache=TRUE}
# Extract actual and predicted values
actual <- test_data_RF$Avg_ridership
predicted <- predict(RF_model, newdata = test_data_RF)

# Create a scatter plot of actual vs. predicted values
plot(actual, predicted, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```

Absolute errors have a significant decrease as well, none, of the bus stops possess an absolute error greater than 75.

```{r, cache=TRUE}
RF_testing_sf <- st_as_sf(RF_testing)

ggplot(RF_testing_sf) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


##### 8.2.2 Generalizability

The residuals are centered around the zero line (marked by the red dashed line), which is generally a good sign. This suggests that there's no systematic bias in the model predictions overall. There is a noticeable increase in the spread of residuals as fitted values increase, particularly visible past a fitted value of about 200. This could indicate heteroscedasticity—where the variability in residuals is not constant across all levels of the predictor variables. In other words, the model might be less reliable at predicting higher values.

```{r, cache=TRUE}
predictions <- predict(RF_model, newdata = train_data_RF)
residuals <- train_data_RF$Avg_ridership - predictions

residual_plot <- ggplot(train_data_RF, aes(x = predictions, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggtitle("Residual Plot of Random Forest Model 1") +
  xlab("Fitted Values") +
  ylab("Residuals")

print(residual_plot)
```


#### 8.3 Random Forest Model 2: Bus Routes
##### 8.3.1 Connecting Stops to Routes

As the transit system is not only formed with stops but connected through routes, and one stop may belong to multiple routes, so it is essential to analyze the pattern of bus routes as well.

```{r, cache=TRUE}

routes_test <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.shp")

new_routes <- routes_test[, c("StopID", "routes")]


# Assuming your dataset is named "new_data" and column "a" contains values like "00040", "00125", etc.
# Convert column "a" to numeric and then format it without leading zeros
new_routes_2 <- new_routes %>%
  mutate(StopID = as.numeric(StopID),
         StopID = sprintf("%d", StopID))


new_routes_2 <- new_routes_2 %>%
  rename('Stop ID' = StopID)


# Convert the sf object "sf_data" to a regular data frame
routes_df <- st_drop_geometry(new_routes_2)

# Remove spaces in "Stop ID" values
data_2022$`Stop ID` <- gsub("\\s+", "", data_2022$`Stop ID`)

# Perform a left join based on the common column "Stop ID"
merged_dataset_bus <- left_join(routes_df, data_2022,   by = "Stop ID")

merged_dataset_bus <- merged_dataset_bus %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


# Separate the routes into individual rows
separated_data <- merged_dataset_bus %>%
  separate_rows(routes, sep = ",\\s*")

# Count the number of routes in each row
# merged_dataset_bus$total_of_routes <- sapply(strsplit(merged_dataset_bus$routes, ",\\s*"), length)


# Calculate the count of unique Stop ID values for each route
# count_stops_data <- separated_data %>% group_by(routes, `Stop ID`) %>% summarise(total_stops_in_routes = n_distinct(`Stop ID`))

count_stops_routes <- separated_data %>%
  group_by(routes) %>%
  summarise(total_stops_in_routes = n_distinct(`Stop ID`))

# Merge the count_stops_data back to the original_data based on routes and counts
merge_count_data <- left_join(separated_data, count_stops_routes, by = "routes")

merge_count_data <- merge_count_data %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


summarized_data2 <- merge_count_data %>%
  group_by(`Stop ID`, Month) %>%
  summarize(total_stops_in_served_routes = sum(total_stops_in_routes, na.rm = TRUE))


# Merge the summarized data back with the original 'bus_2022' dataset
RF2_table <- merge(merged_dataset_bus, summarized_data2, by = c('Stop ID', 'Month'), all.x = TRUE)


RF2_table_sf <- st_as_sf(RF2_table)


```


##### 8.3.2 Running the Model

Same as the random forest model of bus stops, this model has been trained using 500 trees, and at each decision split, 20 variables are randomly considered. Results indicates Mean of Squared Residuals is 10.75963 and a remarkably high 97.03% of the variance explained. 

```{r, cache=TRUE} 

# Split the data into training and testing sets
set.seed(123)  # For reproducibility
train_index2 <- sample(1:nrow(RF2_table_sf), 0.7 * nrow(RF2_table_sf)) 
train_data_RF2 <- RF2_table_sf[train_index2, ]
train_data_RF2 <- na.omit(train_data_RF2)
test_data_RF2 <- RF2_table_sf[-train_index2, ]
test_data_RF2<- na.omit(test_data_RF2)


# Fit the Random Forest model
RF_model3 <- randomForest(
  formula = Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + Month + PopDensity_km2 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center, 
  data = train_data_RF2)


RF_testing3 <- test_data_RF2 %>%
  mutate(Predictions_RF3 = predict(RF_model3, newdata = test_data_RF2),
Error_RF3 = Predictions_RF3 - test_data_RF2$Avg_ridership,
AbsError_RF3 = abs(Predictions_RF3 - test_data_RF2$Avg_ridership),
MAE_RF3= mean(abs(Error_RF3)),
APE_RF3 = (abs(Predictions_RF3 - test_data_RF2$Avg_ridership)) / Predictions_RF3)

RF_model3
```



```{r, cache=TRUE}
# Extract actual and predicted values
actual3 <- test_data_RF2$Avg_ridership
predicted3 <- predict(RF_model3, newdata = test_data_RF2)

# Create a scatter plot of actual vs. predicted values
plot(actual3, predicted3, main = "Actual vs. Predicted", xlab = "Actual", ylab = "Predicted")
abline(0, 1, col = "red")  # Add a line of perfect prediction (y = x)
```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
RF_testing_sf3 <- st_as_sf(RF_testing3)

ggplot(RF_testing_sf3) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF3), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
# Calculating Monthly Prediction
RF_testing_sf_monthly <- RF_testing_sf3 %>%
  mutate(Monthly_prediction = Predictions_RF3 * Number_of_Days)

#file_path <- "predictions.geojson"
#st_write(RF_testing_sf_monthly, file_path, driver = "GeoJSON")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
underperforming_stops_predicted2 <- RF_testing_sf_monthly[RF_testing_sf_monthly$Monthly_prediction <= 10, ]
underperformance_counts_predicted2 <- table(underperforming_stops_predicted2$"Stop ID")

# Top 50 Stops for 2023
top_50_stops_predicted2 <- names(sort(underperformance_counts_predicted2, decreasing = TRUE)[1:50])
underperforming_top_50_predicted2 <- underperforming_stops_predicted2[underperforming_stops_predicted2$"Stop ID" %in% top_50_stops_predicted2, ]

# Order
underperforming_top_50_predicted2$"Stop ID" <- factor(underperforming_top_50_predicted2$"Stop ID", levels = rev(top_50_stops_predicted2))

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, coords = c("Longitude", "Latitude"), crs = 4326)

#underperforming_top_50_sf2 <- st_as_sf(underperforming_top_50_predicted2, wkt = "geometry", crs = 4326)

#underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)




underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_predicted2 <- ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_predicted2, aes(geometry = geometry, colour = q5(Monthly_prediction)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = c("1", "2", "3", "4", "5"),
                     name = "Quintile\nBreaks") +
  labs(title = "50 Bus Stops with Lowest Predicted Avg Ridership",
       fill = "Predictions_RF2") +
  mapTheme()+
  theme(legend.position = "bottom")

print(map_plot_top_50_predicted2)

#file_path <- "Top50underperforming.geojson"
#st_write(underperforming_top_50_predicted2, file_path, driver = "GeoJSON")

```


```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
underperforming_stops_predicted2_m <- underperforming_stops_predicted2[underperforming_stops_predicted2$Month == 'January', ]

ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_stops_predicted2_m, aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January", color = "Predicted Ridership") +
  theme(legend.position = "bottom")

#file_path <- "underperforming_Jan.geojson"
#st_write(underperforming_stops_predicted2_m, file_path, driver = "GeoJSON")
```


##### 8.3.3 Validation

However, to ensure these results are not overly optimistic or specific to the training data, we consider it's crucial to perform validation. So we then set up k-fold cross-validation with 10 folds for training a Random Forest model, ensuring the evaluation is thorough and reliable. This method partitions the data into 10 subsets, uses each subset once as a validation set while training on the other nine, providing a good balance between training size and validation accuracy.

```{r, cache=TRUE}
# Load required packages
install.packages("caret")
library(caret)

# Define the training control parameters for cross-validation
ctrl2 <- trainControl(method = "cv", 
                     number = 10,     # Number of folds
                     verboseIter = TRUE)  

# Train the random forest model using k-fold cross-validation
RF_model2 <- train(Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + Month + PopDensity_km2 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center, 
                  data = train_data_RF2,
                  method = "rf",  
                  trControl = ctrl2)  

# Assess the performance of the model
summary(RF_model2)
```

```{r, cache=TRUE}
RF_model2$resample[1:10,]

mean(RF_model2$resample[,3])
```

The validation result shows the Mean Absolute Error of 1.30 which is comparable goodness of fit metrics across each fold

##### 8.3.4 Generalizability

We saw a similar pattern of the residuals of the random forest model of bus routes and bus stops. The residuals are centered around the zero line. This suggests that there's no systematic bias in the model predictions overall. There is a noticeable increase in the spread of residuals as fitted values increase, particularly visible past a fitted value of about 200. This could indicate heteroscedasticity—where the variability in residuals is not constant across all levels of the predictor variables. The model might be less reliable at predicting higher values as well.
 
```{r, cache=TRUE}
predictions2 <- predict(RF_model2, newdata = train_data_RF2)
residuals2 <- train_data_RF2$Avg_ridership - predictions2

residual_plot <- ggplot(train_data_RF2, aes(x = predictions2, y = residuals2)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  ggtitle("Residual Plot of Random Forest Model 2") +
  xlab("Fitted Values") +
  ylab("Residuals")

print(residual_plot)
```


##### 8.3.5 Tunning Hyperparameters1

Additionally, analyzing which features are most influential in predicting ridership can provide valuable insights and help in refining the model further by adjusting parameters, so we changed the number of trees or the mtry value.

```{r, cache=TRUE}
# Define the parameter grid for tuning only 'mtry'
tuneGrid <- expand.grid(
  mtry = c(10, 15, 20, 25, 30)  # Number of variables randomly sampled as candidates at each split
)
trainControl <- trainControl(
  method = "cv",
  number = 5,  # 5-fold cross-validation
  verboseIter = TRUE
)
# Tune the Random Forest model
set.seed(123)  # Ensure reproducibility
train_index_tuned1 <- sample(1:nrow(RF2_table_sf), 0.7 * nrow(RF2_table_sf)) 
train_data_tuned1 <- RF2_table_sf[train_index_tuned1, ]
train_data_tuned1 <- na.omit(train_data_tuned1)
test_data_tuned1 <- RF2_table_sf[-train_index_tuned1, ]
test_data_tuned1 <- na.omit(test_data_tuned1)

RF_model_tuned1 <- train(
  Avg_ridership ~ TotalPop + MedHHInc + MedRent + Commutetowork + pctWhite + pctAfricanamerican + pctAsian + pctOther + pctMale + pctFemale + pctBachelors + pctWorker + pctPoverty + pctworkwithbus + pctworkfromhome + pctworkbycar + pctworkbytaxi + pctworkbymotor + pctworkbybike + pctworkbywalk + pctnocar + pct1car + pct2car + pct3car + pct4car + pct5car + policeoffice_nn1 + policeoffice_nn2 + policeoffice_nn3 + policeoffice_nn4 + policeoffice_nn5 + groceries_nn1 + groceries_nn2 + groceries_nn3 + groceries_nn4 + groceries_nn5 + parks_nn1 + parks_nn2 + parks_nn3 + parks_nn4 + parks_nn5 + shoppingcen_nn1 + shoppingcen_nn2 + shoppingcen_nn3 + shoppingcen_nn4 + shoppingcen_nn5+ total_routes_in_stops + total_stops_in_served_routes + PopDensity_km2 + transit_stops_nn1 + transit_stops_nn2 + transit_stops_nn3 + transit_stops_nn4 + transit_stops_nn5 + school_nn1 + school_nn2 + school_nn3 + school_nn4 + school_nn5 + dist_to_center,
  data = train_data_tuned1,
  method = "rf",
  metric = "RMSE",
  tuneGrid = tuneGrid,
  trControl = trainControl
)

# Assuming predictions should be made with RF_model_tuned1
predictions_tuned1 <- predict(RF_model_tuned1, test_data_tuned1) 
results_tuned1 <- test_data_RF2 %>%
  mutate(Predictions_tuned1 = predictions_tuned1,
    Error_tuned1 = Predictions_tuned1 - Avg_ridership,
    AbsError_tuned1 = abs(Error_tuned1),
    MAE_tuned1 = mean(AbsError_tuned1),
    APE_tuned1 = (AbsError_tuned1 / Avg_ridership)
  ) %>%
  summarise(MAE = mean(AbsError_tuned1), MAPE = mean(APE_tuned1))

# Print the error metrics
print(results_tuned1)

```

##### 8.3.6 Visualizations for RF Model 2

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
RF_testing_sf2 <- st_as_sf(RF_testing2)

ggplot(RF_testing_sf2) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(aes(color = AbsError_RF2), size = .5) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Bus Stops", color = "Absolute Error Value")

```

We identify and visualize the top 50 underperforming bus stops based on predicted average ridership for 2022. First,we filter out bus stops with predicted ridership values less than or equal to 10 from the RF_testing2 dataset. Then, we rank these underperforming stops based on their frequency of occurrence and selects the top 50. 

```{r, cache=TRUE}
underperforming_stops_predicted2 <- RF_testing2[RF_testing2$Predictions_RF2 <= 10, ]
underperformance_counts_predicted2 <- table(underperforming_stops_predicted2$"Stop ID")

# Top 50 Stops for 2022
top_50_stops_predicted2 <- names(sort(underperformance_counts_predicted2, decreasing = TRUE)[1:50])
underperforming_top_50_predicted2 <- underperforming_stops_predicted2[underperforming_stops_predicted2$"Stop ID" %in% top_50_stops_predicted2, ]

# Order
underperforming_top_50_predicted2$"Stop ID" <- factor(underperforming_top_50_predicted2$"Stop ID", levels = rev(top_50_stops_predicted2))

underperforming_top_50_predicted2$geometry <- st_transform(underperforming_top_50_predicted2$geometry, crs = 3358)  # NAD83 North Carolina

map_plot_top_50_predicted2 <- ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = underperforming_top_50_predicted2, aes(geometry = geometry, colour = q5(Predictions_RF2)), size = 2) +
  scale_color_manual(values = rev(palettebase5),
                     labels = c("1", "2", "3", "4", "5"),
                     name = "Quintile\nBreaks") +
  labs(title = "50 Bus Stops with Lowest Predicted Avg Ridership",
       fill = "Predictions_RF2") +
  mapTheme()+
  theme(legend.position = "bottom")

print(map_plot_top_50_predicted2)
```

We plot a graph that specifically focusing on underperforming stops predicted to have average ridership for the month of January. Most stops are shwoing around 3000 riderships during Janaury,2022.

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
# Calculating Monthly Prediction
RF_testing_sf_monthly <- RF_testing_sf2 %>%
  mutate(Monthly_prediction = Predictions_RF2 * Number_of_Days)

```

```{r, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}
underperforming_stops_predicted2_m <- underperforming_stops_predicted2[underperforming_stops_predicted2$Month == 'January', ]

ggplot() +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = RF_testing_sf_monthly, aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership in January", color = "Predicted Ridership") +
  theme(legend.position = "bottom")

```

##### 8.3.7 Multicolinearity

Variables such as transit_stops_nn5, transit_stops_nn4, total_stops_in_served_routes, and groceries_nn1 show very high importance scores, indicating they are significant predictors in determining average ridership. The high scores for proximity to transit stops and groceries suggest that accessibility and local amenities are major factors affecting ridership.
Demographic factors like pctAfricanamerican, pctFemale, MedHHInc, and population-related variables like TotalPop have moderate importance. This indicates that community demographics and economic factors also play a significant role in predicting ridership.
Variables such as pctworkbybike, pctworkbymotor, and some of the nearest neighbor variables for schools have lower importance scores, suggesting they have less impact on ridership levels compared to others.
The visualization of these importance scores would rank these variables by their scores, providing a clear visual hierarchy of feature significance. 

```{r, cache=TRUE}
# Load the necessary libraries
library(randomForest)

# Assuming RF_model2 is a trained model using caret's train function
# Extract the final model (if it's indeed a Random Forest model)
final_model <- RF_model2$finalModel

# Get variable importance scores from the Random Forest model
importance_scores <- importance(final_model)

# Print the variable importance scores
print(importance_scores)

# Plot the variable importance scores
varImpPlot(final_model)
```


### 9. Use-Case Scenario

#### 9.1 Scenario 1: Increase in Population Density

##### 9.1.1 Increase in Population Density 15% (2022 & 2027)

The last measured population density for Charlotte, NC was 2,827 in 2018. Charlotte, NC experienced an average growth rate of 2.87% from our first statistic recorded in 2009. If past trends continue, we forecast the population density growth rate would be 15% from 2022 to 2027 and 30% from 2022 to 2032.
Population density increase does not affect average ridership by much but decreases underperforming stops. In both 2027 and 2032, we would see ridership larger than 300 at stops located at the southwestern periphery and downtown of the county, but still the majority of the stops with ridership under 100. The southeastern and northern area of the county is having the most stops with average ridership under 10 in January 2027 and January 2032, but the total number of stops with average ridership decresed from 2027 to 2032.


```{r, scenario 1 2022, warning=FALSE, message=FALSE, fig.height=8, fig.width=16, cache=TRUE}

# Simulate 15% increase in PopDensity
RF2_table_2022_scenario1 <- RF2_table_sf %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.15)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario1 <- predict(RF_model3, newdata = RF2_table_2022_scenario1)

# Add Predictions to the Data
RF2_table_2022_scenario1$Predicted_Avg_ridership <- predictions2_2022_scenario1

# Add Monthly Prediction Column
RF2_table_2022_scenario1 <- RF2_table_2022_scenario1 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
#RF2_table_2022_scenario1$Monthly_prediction <- sprintf("%.2f", predictions2_2022_scenario1)

# Convert Data to an sf Object
RF2_table_2022_scenario1_sf <- st_as_sf(RF2_table_2022_scenario1, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario1_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario1_sf$Monthly_prediction)

result <- RF2_table_2022_scenario1_sf[RF2_table_2022_scenario1_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario I", color = "Predicted Avg Ridership with \n15% Increase in Population Density")+
  theme(legend.position = "bottom")

```


##### 9.1.2 Increase in Population Density 30%


```{r, scenario 1a 2022, cache=TRUE}

# Simulate 30% increase in PopDensity
RF2_table_2022_scenario1a <- RF2_table_sf %>% 
  mutate(PopDensity_km2 = PopDensity_km2 * 1.3)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario1a <- predict(RF_model3, newdata = RF2_table_2022_scenario1a)

# Add Predictions to the Data
RF2_table_2022_scenario1a$Predicted_Avg_ridership <- predictions2_2022_scenario1a

# Add Monthly Prediction Column
RF2_table_2022_scenario1a <- RF2_table_2022_scenario1a %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Convert Data to an sf Object
RF2_table_2022_scenario1a_sf <- st_as_sf(RF2_table_2022_scenario1a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario1a_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario1a_sf$Monthly_prediction)


result <- RF2_table_2022_scenario1a_sf[RF2_table_2022_scenario1a_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario I", color = "Predicted Avg Ridership with \n30% Increase in Population Density")+
  theme(legend.position = "bottom")
```


#### 9.2 Scenario 2: Increase in Median Income

##### 9.2.1 Increase in Income 25%

Between 2020 and 2021 Charlotte's median household income grew from $65,359 to $68,367, a 4.6% increase. It will be an increase of 25% from 2022 to 2027 and 50% from2022 to 2032.
As income increases, the average ridership does not change much. Average ridership larger than 300 at stops located at the southwestern periphery and downtown of the county, but still the majority of the stops with ridership under 100. However, less bus stops is having monthly average ridership under 10 in 2032 than 2027.


```{r, scenario 2 2022, cache=TRUE}

RF2_table_2022_scenario2 <- RF2_table_sf %>% 
  mutate(MedHHInc = MedHHInc * 1.25)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario2 <- predict(RF_model3, newdata = RF2_table_2022_scenario2)

# Add Predictions to the Data
RF2_table_2022_scenario2$Predicted_Avg_ridership <- predictions2_2022_scenario2

# Add Monthly Prediction Column
RF2_table_2022_scenario2 <- RF2_table_2022_scenario2 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario2$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario2)

# Convert Data to an sf Object
RF2_table_2022_scenario2_sf <- st_as_sf(RF2_table_2022_scenario2, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario2_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario2_sf$Monthly_prediction)


result <- RF2_table_2022_scenario2_sf[RF2_table_2022_scenario2_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario II", color = "Predicted Avg Ridership with \n25% Increase in Median Income")+
  theme(legend.position = "bottom")

```


##### 9.2.2 Increase in Median Income 50%

```{r, scenario 2a 2022, cache=TRUE}

# Simulate 10% increase in PopDensity
RF2_table_2022_scenario2a <- RF2_table_sf %>% 
  mutate(MedHHInc = MedHHInc * 1.5)

# Make Predictions with the New Scenario Data
predictions2_2022_scenario2a <- predict(RF_model3, newdata = RF2_table_2022_scenario2a)

# Add Predictions to the Data
RF2_table_2022_scenario2a$Predicted_Avg_ridership <- predictions2_2022_scenario2a

# Add Monthly Prediction Column
RF2_table_2022_scenario2a <- RF2_table_2022_scenario2a %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Format Predicted Avg Ridership as Decimal Numbers
# RF2_table_2022_scenario2a$Predicted_Avg_ridership <- sprintf("%.2f", predictions2_2022_scenario2a)

# Convert Data to an sf Object
RF2_table_2022_scenario2a_sf <- st_as_sf(RF2_table_2022_scenario2a, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario2a_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario2a_sf$Monthly_prediction)


result <- RF2_table_2022_scenario2a_sf[RF2_table_2022_scenario2a_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario II", color = "Predicted Avg Ridership with \n50% Increase in Median Income")+
  theme(legend.position = "bottom")

```


#### 9.3 Scenario 3: Add 2 Proposed Light Rail Stops as Feature

The density of new stops of the proposed light rail is the highest in the city center area.
Following the same process of modeling and visualizing, we test the dataset adding the two proposed light rail data using random forest and assessed its performance using Actual vs. Prediction plot and absolute error map. We also mapping out the underperforming stops.

```{r, fig.height=8, fig.width=16, cache=TRUE}
# Rename OBJECTID_1 column to OBJECTID in red_stops
red_stops <- red_stops %>% rename(OBJECTID = OBJECTID_1)
red_stops <- red_stops %>% rename(NAME = Name)
red_stops <- select(red_stops, -Status)


# Add a unique identifier to silver_stops
silver_stops <- select(silver_stops, -c(Phase, parkRide, GeoID))
silver_stops <- silver_stops %>% rename(NAME = Name)

# Merge the datasets based on OBJECTID
proposed_line_merged_data <- rbind(existing_transit_stops, red_stops)
proposed_line_merged_data <- rbind(proposed_line_merged_data, silver_stops)

# Assuming your dataset is named "stops_data"
proposed_line_merged_data_sf <- st_as_sf(proposed_line_merged_data, coords = c("geometry.x", "geometry.y"))


# Plot the density of points

ggplot() +
  geom_sf(data = dat2020, fill = "grey90", alpha = 0.5, color = NA) +
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = proposed_line_merged_data, color = "black", size = .5, alpha = 0.5) +
  stat_density_2d(data = proposed_line_merged_data_sf, 
                  aes(x = st_coordinates(geometry)[, 1], 
                      y = st_coordinates(geometry)[, 2],
                      fill = ..level.., alpha = ..level..),
                  size = 0.01, bins = 40, geom = 'polygon') +
  scale_fill_gradient(low = "#737373", high = "#0077B6", name = "Density") +  # Adjust color gradient
  scale_alpha(range = c(0.00, 0.1), guide = "none") +
  labs(title = "Density of Proposed Lines") +
  mapTheme()
  
```


```{r, cache=TRUE}
RF2_table_sf <- na.omit(RF2_table_sf)

data_2022_s1 <-
  RF2_table_sf %>% 
    mutate(
      proposed_line_nn1 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 1),
      
      proposed_line_nn2 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 2), 
      
      proposed_line_nn3 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 3), 
      
      proposed_line_nn4 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 4), 
      
      proposed_line_nn5 = nn_function(st_coordinates(RF2_table_sf), 
                              st_coordinates(proposed_line_merged_data_sf), k = 5)) 

```


```{r, cache=TRUE}
data_2022_s1 <- subset(data_2022_s1, select = -c(transit_stops_nn1, transit_stops_nn2, transit_stops_nn3, transit_stops_nn4, transit_stops_nn5))

# Rename columns using base R
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn1"] <- "transit_stops_nn1"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn2"] <- "transit_stops_nn2"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn3"] <- "transit_stops_nn3"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn4"] <- "transit_stops_nn4"
names(data_2022_s1)[names(data_2022_s1) == "proposed_line_nn5"] <- "transit_stops_nn5"

```


```{r, scenario 2a 2022, cache=TRUE}


# Make Predictions with the New Scenario Data
predictions2_2022_scenario3 <- predict(RF_model3, newdata = data_2022_s1)

# Add Predictions to the Data
data_2022_s1$Predicted_Avg_ridership <- predictions2_2022_scenario3

# Add Monthly Prediction Column
data_2022_s1 <- data_2022_s1 %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_Days)

# Convert Data to an sf Object
RF2_table_2022_scenario3_sf <- st_as_sf(data_2022_s1, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
RF2_table_2022_scenario3_sf$Monthly_prediction <- as.numeric(RF2_table_2022_scenario3_sf$Monthly_prediction)


result <- RF2_table_2022_scenario3_sf[RF2_table_2022_scenario3_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = silver_routes, color = "purple", size = 7, alpha = 0.5) +
  geom_sf(data = red_routes,color = "#FF6347", size = 7, alpha = 0.5) +
  geom_sf(data = blue_routes, color = "#4682B4", size = 7, alpha = 0.5) +
  geom_sf(data = gold_routes,color = "#FFD700", size = 7, alpha = 0.5) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario III",
       color = "Predicted Avg Ridership \nwith Proposed Transit Lines Added") +
  theme(
    legend.position = "bottom",
    plot.title = element_text(size = 10)  # Adjust size here to your preference
  )

```


#### 9.4 Senario 4: Test add 10 stops

We add 10 stops to the original dataset to see what will happen when new stops are built. Again following the same step using Actual vs.Prediction scatterplot and map of absolute errors to visualize the performance, also mapping out the underperforming stops when 10 more stops are added.


```{r, cache=TRUE}
newstops_test <- st_read("Test/2022_test.shp")%>%
  st_transform('EPSG:3358')
```


```{r, warning=FALSE, message=FALSE, cache=TRUE}
# Perform spatial join based on containment (bus stops within census tracts)
joined_data_test <- st_join(newstops_test, dat2022, join = st_within)
```


```{r, cache=TRUE}
joined_data_test <-
  joined_data_test %>% 
    mutate(
      school_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(school.sf), k = 1),
      
      school_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(school.sf), k = 2), 
      
      school_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(school.sf), k = 3), 
      
      school_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(school.sf), k = 4), 
      
      school_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(school.sf), k = 5)) 

joined_data_test <-
  joined_data_test %>% 
    mutate(
      policeoffice_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(policeoffice.sf), k = 1),
      
      policeoffice_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(policeoffice.sf), k = 2), 
      
      policeoffice_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(policeoffice.sf), k = 3), 
      
      policeoffice_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(policeoffice.sf), k = 4), 
      
      policeoffice_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(policeoffice.sf), k = 5)) 

joined_data_test <-
  joined_data_test %>% 
    mutate(
      groceries_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(groceries.sf), k = 1),
      
      groceries_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(groceries.sf), k = 2), 
      
      groceries_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(groceries.sf), k = 3), 
      
      groceries_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(groceries.sf), k = 4), 
      
      groceries_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(groceries.sf), k = 5)) 


joined_data_test <-
  joined_data_test %>% 
    mutate(
      parks_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(parks.sf), k = 1),
      
      parks_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(parks.sf), k = 2), 
      
      parks_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(parks.sf), k = 3), 
      
      parks_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(parks.sf), k = 4), 
      
      parks_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(parks.sf), k = 5)) 

joined_data_test <-
  joined_data_test %>% 
    mutate(
      shoppingcen_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(shoppingcen.sf), k = 1),
      
      shoppingcen_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(shoppingcen.sf), k = 2), 
      
      shoppingcen_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(shoppingcen.sf), k = 3), 
      
      shoppingcen_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(shoppingcen.sf), k = 4), 
      
      shoppingcen_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(shoppingcen.sf), k = 5)) 

```


```{r, cache=TRUE}
joined_data_test <-
  joined_data_test %>% 
    mutate(
      transit_stops_nn1 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(existing_transit_stops), k = 1),
      
      transit_stops_nn2 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(existing_transit_stops), k = 2), 
      
      transit_stops_nn3 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(existing_transit_stops), k = 3), 
      
      transit_stops_nn4 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(existing_transit_stops), k = 4), 
      
      transit_stops_nn5 = nn_function(st_coordinates(joined_data_test), 
                              st_coordinates(existing_transit_stops), k = 5)) 
```


```{r, cache=TRUE}
city_center<- st_sfc(st_point(c(1692115, 5236081)), crs = 3358)
joined_data_test$dist_to_center <- st_distance(joined_data_test$geometry, city_center)
```

```{r, cache=TRUE}
joined_data_test <- joined_data_test %>% 
             mutate(Avg_ridership = Avg_Board + Avg_Alight)
```


```{r, cache=TRUE}
routes_test <- st_read("Bus Transit Data/Bus_Stops_With_Frequency_HLT.geojson")

new_routes <- routes_test[, c("StopID", "routes")]


# Assuming your dataset is named "new_data" and column "a" contains values like "00040", "00125", etc.
# Convert column "a" to numeric and then format it without leading zeros
new_routes_2 <- new_routes %>%
  mutate(StopID = as.numeric(StopID),
         StopID = sprintf("%d", StopID))


new_routes_2 <- new_routes_2 %>%
  rename('Stop ID' = StopID)

# Convert the sf object "sf_data" to a regular data frame
routes_df <- st_drop_geometry(new_routes_2)

# Remove spaces in "Stop ID" values
joined_data_test$`Stop ID` <- gsub("\\s+", "", joined_data_test$`Stop.ID`)

# Perform a left join based on the common column "Stop ID"
merged_dataset_bus2 <- left_join( routes_df,joined_data_test, by = "Stop ID")

merged_dataset_bus2 <- merged_dataset_bus2 %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


# Separate the routes into individual rows
separated_data2 <- merged_dataset_bus2 %>%
  separate_rows(routes, sep = ",\\s*")

count_stops_routes2 <- separated_data2 %>%
  group_by(routes) %>%
  summarise(total_stops_in_routes = n_distinct(`Stop ID`))

# Merge the count_stops_data back to the original_data based on routes and counts
merge_count_data2 <- left_join(separated_data2, count_stops_routes2, by = "routes")

merge_count_data2 <- merge_count_data2 %>%
  mutate("total_routes_in_stops" = str_count(routes, ",") + 1)


summarized_data22 <- merge_count_data2 %>%
  group_by(`Stop ID`, Month) %>%
  summarize(total_stops_in_served_routes = sum(total_stops_in_routes, na.rm = TRUE))


# Merge the summarized data back with the original 'bus_2022' dataset
RF2_table2 <- merge(merged_dataset_bus2, summarized_data22, by = c('Stop ID', 'Month'), all.x = TRUE)


RF2_table2_sf <- st_as_sf(RF2_table2)


```

```{r, cache=TRUE}
# Extract rows from table2 where stop_name matches specific values
filtered_table2 <- joined_data_test[joined_data_test$Stop.Name %in% c("Test_1", "Test_2", "Test_3", "Test_4", "Test_5", "Test_6", "Test_7", "Test_8", "Test_9", "Test_10"), ]

filtered_table2$routes <- 1
filtered_table2$total_routes_in_stops <- 1
filtered_table2$total_stops_in_served_routes <- 1

colnames(RF2_table2)<-colnames(filtered_table2)  

table_test <- rbind(RF2_table2_sf, filtered_table2)
```


```{r, cache=TRUE}
# Make Predictions with the New Scenario Data
predictions_add_stops <- predict(RF_model3, newdata = table_test)

# Add Predictions to the Data
table_test$Predicted_Avg_ridership <- predictions_add_stops

# Add Monthly Prediction Column
table_test <- table_test %>%
  mutate(Monthly_prediction = Predicted_Avg_ridership * Number_of_)

# Convert Data to an sf Object
table_test_sf <- st_as_sf(table_test, geom = "geometry")


# Convert Predicted Avg Ridership to Numeric
table_test_sf$Monthly_prediction <- as.numeric(table_test_sf$Monthly_prediction)

result <- table_test_sf[table_test_sf$Monthly_prediction <= 10, ]
result <- result[result$Month == 'January', ]

filtered_table2_sf <- st_as_sf(filtered_table2, geom = "geometry")

# Plot the Map with Correct Fill Scale
ggplot(result) +
  geom_sf(data = allTracts_sf, fill = "grey90", alpha = 0.8 , color = NA)+
#  geom_sf(data = dat2022,aes(fill = (PopDensity_km2)), alpha = 0.5) +
#  scale_fill_gradient(low = "white", high = "grey20") + 
  geom_sf(data = bus_routes, color = "darkgrey", size = 5, alpha = 0.5) +
  geom_sf(data = filtered_table2_sf, fill = "red", alpha = 0.8) +
  geom_sf(aes(color = Monthly_prediction), size = 1) +
  scale_color_gradient(low = "#FFBD37", high = "#005357") +
  mapTheme() +
  labs(title = "Predicted Avg Ridership under 10 in January: Scenario IV", color = "Predicted Avg Ridership with \n10 New Added Stops")+
  theme(legend.position = "bottom",
    plot.title = element_text(size = 10)  # Adjust size here to your preference
  )

```









