---
title: "Census Data"
author: " "
date: '2023-09-08'
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load packages and functions

```{r setup_packages, warning = FALSE, message = FALSE}
# Load Libraries

library(tidyverse)
library(tidycensus)
library(sf)
library(kableExtra)

options(scipen=999)
options(tigris_class = "sf")

source("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/functions.r")

palette5 <- c("#f0f9e8","#bae4bc","#7bccc4","#43a2ca","#0868ac", "#001f3f")
```

```{r load_key, warning = FALSE, eval = FALSE}
census_api_key("67608da2d7728d1edc9cb636b83bb764101fe726", overwrite = TRUE)
```

```{r setup_packages, warning = FALSE, message = FALSE}
# Create a vector of years
years <- c(2017, 2018, 2019, 2020, 2021, 2022)

# Loop through the years to load variables
variables_list <- lapply(years, function(year) {
  load_variables(year, "acs5", cache = TRUE)
})

# Combine the variables from different years into a single data frame
all_variables <- do.call(rbind, variables_list)
```


### 2017 Census Data

```{r results='hide'}
tracts17 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2017, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2017") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2018 Census Data

```{r results='hide'}
tracts18 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2018, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2018") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2019 Census Data

```{r results='hide'}
tracts19 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2019, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2019") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2020 Census Data

```{r results='hide'}
tracts20 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2020, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2020") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2021 Census Data

```{r results='hide'}
tracts21 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2021, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2021") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### 2022 Census Data

```{r results='hide'}
tracts22 <- 
  get_acs(geography = "tract", 
          variables = c("B25026_001E","B02001_002E",
                        "B02001_003E","B02001_005E",
                        "B15001_050E","B15001_009E",
                        "B19013_001E","B25058_001E",
                        "B06012_002E","B08301_001E",
                        "B08301_011E","B01001_002E",
                        "B01001_026E","B08111_031E",
                        "B24080_001E","B24080_002E",
                        "B24080_012E"), 
          year=2022, state=37, county=119, 
          geometry=TRUE, output="wide") %>%
  st_transform('ESRI:102728') %>%
  rename(TotalPop = B25026_001E, 
         Whites = B02001_002E,
         Africanamerican = B02001_003E,
         Asian = B02001_005E,
         Male = B01001_002E,
         Female = B01001_026E,
         FemaleBachelors = B15001_050E, 
         MaleBachelors = B15001_009E,
         Workermale= B24080_002E,
         Workerfemale= B24080_012E,
         MedHHInc = B19013_001E, 
         MedRent = B25058_001E,
         TotalPoverty = B06012_002E,
         Transporttowork = B08301_001E,
         Workfromhome = B08111_031E,
         Workwithbus = B08301_011E) %>%
  dplyr::select(-NAME, -starts_with("B")) %>%
  mutate(TotalWorker = (Workermale +Workerfemale),
         pctWhite = ifelse(TotalPop > 0, Whites / TotalPop,0),
         pctAfricanamerican = ifelse(TotalPop > 0, Africanamerican / TotalPop,0),
         pctAsian = ifelse(TotalPop > 0, Asian / TotalPop,0),
         pctOther = ifelse(TotalPop > 0, (TotalPop - Whites - Africanamerican - Asian) / TotalPop,0),
         pctMale = ifelse(TotalPop > 0, (Male / TotalPop),0),
         pctFemale = ifelse(TotalPop > 0, (Female / TotalPop),0),
         pctBachelors = ifelse(TotalPop > 0, ((FemaleBachelors + MaleBachelors) / TotalPop),0),
         pctWorker = ifelse(TotalPop > 0, (Workermale + Workerfemale) / TotalPop, 0),
         pctPoverty = ifelse(TotalPop > 0, TotalPoverty / TotalPop, 0),
         pctworkwithbus = ifelse(Transporttowork > 0, Workwithbus / Transporttowork, 0),
         pctworkfromhome = ifelse(Transporttowork > 0, Workfromhome / Transporttowork, 0),
         year = "2022") %>%
  dplyr::select(-Whites, -Africanamerican, -Asian, -Male, -Female, -FemaleBachelors, -MaleBachelors, -TotalPoverty, -Workermale, -Workerfemale) 
```


### Combining Census Data from 2017 - 2022

```{r}

allTracts <- rbind(tracts17,tracts18,tracts19,tracts20,tracts21,tracts22)
```


### Creating geojson

```{r}
sf_allTracts <- st_as_sf(allTracts, wkt = "geometry")
```


```{r}

total_population_per_year <- allTracts %>%
  group_by(year) %>%
  summarize(TotalPopulation = sum(TotalPop))

print(total_population_per_year)
```

### Census Data Trend

```{r Trends by Year, warning=FALSE, fig.width=8, fig.height=8}

a <- ggplot(allTracts, aes(x = as.factor(year), y = TotalPop, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Total Population: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palette5)

b <- ggplot(allTracts, aes(x = as.factor(year), y = TotalWorker, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Worker: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palette5)

c <- ggplot(allTracts, aes(x = as.factor(year), y = Workfromhome, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of Work from Home: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palette5)

d <- ggplot(allTracts, aes(x = as.factor(year), y = Workwithbus, fill = as.factor(year))) +
  geom_bar(stat = "identity", color = "transparent") +
  labs(title = "Population of People Using Bus as Transportation to Work: 2017-2022", x = "Year", y = "Total Population") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(legend.position = "none") +
  theme(plot.title = element_text(size = 6)) +
  scale_fill_manual(values = palette5)

combined_plot <- grid.arrange(a,b,c, d, ncol = 2, top = c(2, 2))

print(combined_plot)

```


### Bus Transit Data

```{r Load Bus Transit Data, warning = FALSE, eval = FALSE, results='hide', message=FALSE}

bus_data <- read.csv("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/AllDat.csv")

routes <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Bus_Routes.geojson")

hlt <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/Bus_Stops_With_Frequency_HLT.geojson")

CATSstop <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/CATS_BusStops/CATS_BusStops.shp")

census <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/combined.geojson")

boundary <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/MecklenburgCounty_Boundary/MecklenburgCounty_Boundary.shp")

```

```{r data wrangling, warning = FALSE, eval = FALSE, results='hide'}
bus_data <- bus_data %>% mutate(StopID = sprintf("%05d", Stop.ID))
```

### Projection

```{r projection, warning = FALSE, eval = FALSE, results='hide'}
print("CRS of busstop_data:")
print(st_crs(bus_data))
print("CRS of catsstop_data:")
print(st_crs(CATSstop))
print("CRS of routes:")
print(st_crs(routes))
```


```{r transform, warning = FALSE, eval = FALSE, results='hide'}
# Create simple feature (sf) objects with initial CRS
busstop_sf <- st_as_sf(bus_data, coords = c("Longitude", "Latitude"), crs = 4326)
catsstop_sf <- st_as_sf(CATSstop, coords = c("LONGITUDE", "LATITUDE"), crs = 2264)  # Assuming NAD 83/North Carolina (ftUS)

# Reproject catsstop_sf to WGS 84 (EPSG:4326)
catsstop_sf <- st_transform(catsstop_sf, 4326)
boundary_sf <- st_transform(boundary, 4326)
```

```{r transformed projection, warning = FALSE, eval = FALSE, results='hide'}
print("CRS of busstop_data:")
print(st_crs(busstop_sf))
print("CRS of catsstop_data:")
print(st_crs(catsstop_sf))
print("CRS of routes:")
print(st_crs(routes))
print("CRS of boundary:")
print(st_crs(boundary_sf))
```


### Routes and Stops in Mecklenburg County

```{r Routes and Stops, warning = FALSE, eval = FALSE, results='hide'}

ggplot() +
  geom_sf(data = routes, aes(color = ROUTE_NAME), size = 2) +
  geom_sf(data = CATSstop, color = "red", size = .2) +
  geom_sf(data = boundary_sf, fill = NA, color = "black", size = 1) +
  
  
  labs(title = "Bus Routes and Stops in Mecklenburg County") +
  
  scale_color_viridis_d() +  
  
  theme_minimal() +
  theme(
    legend.position = "none"  
  )
```


### CATS Bus Routes

```{r plot routes and stops, warning = FALSE, eval = FALSE, results='hide'}
plots_list <- list()

for (route_name in unique(routes$ROUTE_NAME)) {
  
  subset_routes <- subset(routes, ROUTE_NAME == route_name)
  subset_stops <- subset(CATSstop, routes == route_name)
  
  p <- ggplot() +
    geom_sf(data = subset_routes, aes(color = ROUTE_NAME), size = 2) +
    geom_sf(data = subset_stops, color = "red", size = .8) +
    geom_sf(data = boundary_sf, fill = NA, color = "black", size = 1) +
    labs(title = paste("Bus Route and Stops -", route_name)) +
    scale_color_viridis_d() +
    theme_minimal() +
    theme(legend.position = "none")
  
  plots_list[[as.character(route_name)]] <- p
}

for (i in seq_along(plots_list)) {
  print(plots_list[[i]])
}
```


### APC data

```{r Board, warning = FALSE, eval = FALSE, results='hide'}
ggplot(busstop_sf) +
  geom_sf(aes(color = Board), size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Bus Stops", color = "Board Value")
```


```{r Alight, warning = FALSE, eval = FALSE, results='hide'}
ggplot(busstop_sf) +
  geom_sf(aes(color = Alight), size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Bus Stops", color = "Alight Value")
```


### Underperform Bus Stop

```{r underperform stops, warning = FALSE, eval = FALSE, results='hide'}
filtered_data <- busstop_sf %>% filter(Board <= 5, Alight <= 5)

filtered_plot <- ggplot(filtered_data) +
  geom_sf(aes(color = Board + Alight), size = 1) +
  scale_color_gradient(low = "blue", high = "red") +
  theme_minimal() +
  labs(title = "Bus Stops with Board and Alight Value <= 5", color = "Combined Value")

print(filtered_plot)
```

### Number of Underperform Bus Stops

```{r number of underperform stops, warning = FALSE, eval = FALSE, results='hide'}
unique_stop_ids <- filtered_data %>%
  summarise(unique_stop_ids = n_distinct(Stop.ID))

print(unique_stop_ids$unique_stop_ids)
```

### Frequency of Underperform Bus Stops

```{r frequency of underperform stops, warning = FALSE, eval = FALSE, results='hide'}
top10_bus_stops <- filtered_data %>%
  group_by(Stop.ID) %>%
  summarise(stop_count = n()) %>%
  arrange(desc(stop_count)) %>%
  head(10)

ggplot(top10_bus_stops, aes(x = factor(Stop.ID), y = stop_count)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(title = "Top 10 Bus Stops by Frequency", x = "Bus Stop ID", y = "Frequency") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


```{r underperform stops for years, warning = FALSE, eval = FALSE, results='hide'}
top5_bus_stops_per_year <- filtered_data %>%
  filter(Year >= 2017 & Year <= 2022) %>%
  group_by(Year, Stop.ID) %>%
  summarise(combined_value = sum(Board + Alight)) %>%
  arrange(Year, desc(combined_value)) %>%
  group_by(Year) %>%
  slice_head(n = 5)

ggplot(top5_bus_stops_per_year, aes(x = factor(Stop.ID), y = combined_value, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = "dodge", color = "white") +
  scale_fill_manual(values = palette5) +
  labs(title = "Top 5 Bus Stops by Combined Board and Alight Value (2017-2022)", x = "Bus Stop ID", y = "Combined Value", fill = "Year") +
   theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~Year, scales = "free_x", ncol = 6)
```


### Underperform Bus Stops per Years

```{r plot of underperform stops for years, warning = FALSE, eval = FALSE, results='hide'}

top5_bus_stops_per_year <- filtered_data %>%
  filter(Year >= 2017 & Year <= 2022) %>%
  group_by(Year, Stop.ID) %>%
  summarise(combined_value = sum(Board + Alight)) %>%
  arrange(Year, desc(combined_value)) %>%
  group_by(Year) %>%
  slice_head(n = 5)

maps_list <- lapply(unique(top5_bus_stops_per_year$Year), function(year) {
  year_data <- filter(top5_bus_stops_per_year, Year == year)
  
  ggplot() +
    geom_sf(data = busstop_sf, color = "gray", size = 0.5) +
    geom_sf(data = year_data, aes(color = combined_value), size = 2) +
    scale_color_viridis_c() + 
    labs(title = paste("Top 5 Bus Stops in", year),
         subtitle = "Combined Board and Alight Value",
         color = "Combined Value") +
    theme_minimal()
})

for (i in seq_along(maps_list)) {
  print(maps_list[[i]])
}

```


```{r least bus stop with combined Board and Alight per year, warning = FALSE, eval = FALSE, results='hide'}
least5_bus_stops_per_year <- busstop_sf %>%
  filter(Year >= 2017 & Year <= 2022) %>%
  group_by(Year, Stop.ID) %>%
  summarise(combined_value = sum(Board + Alight)) %>%
  arrange(Year, desc(combined_value)) %>%
  group_by(Year) %>%
  slice_tail(n = 5)

ggplot(least5_bus_stops_per_year, aes(x = factor(Stop.ID), y = combined_value, fill = as.factor(Year))) +
  geom_bar(stat = "identity", position = "dodge", color = "white") +
  scale_fill_manual(values = palette5) +  
  labs(title = "Least 5 Bus Stops by Combined Board and Alight Value (2017-2022)", x = "Bus Stop ID", y = "Combined Value", fill = "Year") +
   theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~Year, scales = "free_x", ncol = 6)
```


```{r plot of least bus stop with combined Board and Alight per year, warning = FALSE, eval = FALSE, results='hide'}

maps_list <- lapply(unique(least5_bus_stops_per_year$Year), function(year) {
  year_data <- filter(least5_bus_stops_per_year, Year == year)
  
  ggplot() +
    geom_sf(data = busstop_sf, color = "gray", size = 0.5) +
    geom_sf(data = year_data, aes(color = combined_value), size = 2) +
    scale_color_viridis_c() + 
    labs(title = paste("Least 5 Bus Stops in", year),
         subtitle = "Combined Board and Alight Value",
         color = "Combined Value") +
    theme_minimal()
})

for (i in seq_along(maps_list)) {
  print(maps_list[[i]])
}

```










































### Relating CATS Stops and Tracts
#### Wrangling Charlotte Transit Open Data

```{r}

routes <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Bus_Routes.geojson")

hlt <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/Census/Bus_Stops_With_Frequency_HLT.geojson")

CATSstop <- st_read("C:/Users/jonat/OneDrive - PennO365/Documents/MUSA 8010/Practicum_Charlotte_2024/CATS_BusStops/CATS_BusStops.shp")
```


```{r}

stopBuffer <- st_buffer(CATSstop, 2640)

stopUnion <- st_union(st_buffer(CATSstop, 2640))

CATSBuffer <- 
  rbind(
     stopBuffer %>%
      mutate(Legend = "Buffer") %>%
      dplyr::select(Legend),
     stopUnion %>%
      st_sf() %>%
      mutate(Legend = "Unioned Buffer"))
```


```{r}
ggplot() +
  geom_sf(data=CATSBuffer) +
  geom_sf(data=CATSstop, show.legend = "point") +
  facet_wrap(~Legend) + 
  labs(caption = "Figure 1") +
  mapTheme()
```


#### Using the `sf` Package for Spatial operations

```{r}
buffer <- filter(CATSBuffer, Legend=="Unioned Buffer")

buffer <- st_transform(buffer, st_crs(tracts17))
```


#### Spatial Intersection of tracts and buffers

##### Spatial Clipping with `st_intersection()` on polygons

```{r}
clip <- 
  st_intersection(buffer, tracts17) %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Clip")
```


##### Spatial selection

```{r spatialSelection}

# Do a spatial selection to see which tracts touch the buffer

# approach #1: sub-setting a spatial object with a spatial object using the '[' brackets.
selection1 <- 
  tracts17[buffer,] %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Spatial Selection")

```


##### Spatial intersection with with `st_centroid()` on polygon centroids

```{r}
selectCentroids <-
  st_centroid(tracts17)[buffer,] %>%
  st_drop_geometry() %>%
  left_join(., dplyr::select(tracts17, GEOID), by = "GEOID") %>%
  st_sf() %>%
  dplyr::select(Workwithbus) %>%
  mutate(Selection_Type = "Select by Centroids")
```

##### Plot

```{r}
intersections <- rbind(clip, selection1, selectCentroids)

ggplot() +
  geom_sf(data=intersections, aes(fill = Workwithbus)) +
  geom_sf(data=CATSstop, show.legend = "point", size = 0.7) +
  scale_fill_viridis_c() +
  facet_wrap(~Selection_Type) + 
  mapTheme()
```

